---
title: "Comparing execution time with varying processors"
author: "Diana Betelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing execution time with varying processors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**Introduction**

In this vignette we will analyse how the number of cores used affect the execution
time of the Pscan algorithm. We will run the algorithm with 1, 2, 4, 6, 8, 12, 18, and 24
cores.

In this experiment we will use the biggest dataset aviable, the liver one, with promoter 
sequences length equal to 1000 bp. 
Notice that the observed speedup will depend on the dimension of the data.

```{r}
file_path <- system.file("extdata", "df_processori.rds", package = "PscanR")
data<- readRDS(file_path)

data <- t(data)

N <- as.numeric(colnames(data))
means <- colMeans(data, na.rm = TRUE)

# Power regression
power_model <- lm(log(means) ~ log(N))  

smooth_N <- seq(min(N), max(N), length.out = 500)  # 500 punti tra 1 e 24
smooth_pred <- exp(predict(power_model, newdata = data.frame(N = smooth_N)))  # Previsioni

ss_total <- sum((means - mean(means))^2)
ss_residual <- sum((means - exp(predict(power_model)))^2)
r_squared <- 1 - (ss_residual / ss_total) 

```

```{r boxplot-example, fig.width=6, fig.height=4}
boxplot(data,
        main = "PscanR Execution Time",
        xlab = "Cores (N)\nLiver dataset (950d−50u), JASPAR2020 (746 matrices), 329 promoters (1000bp)",
        cex.lab = 0.8,
        ylab = "Time (s)",
        col = "lightblue",
        las = 1, names = N, outline = FALSE)

lines(smooth_N, smooth_pred, col = "blue", lwd = 2)

legend("topright", legend = bquote("Power regression, " ~ R^2 == .(round(r_squared, 3))),
       col = "blue", lwd = 2)
```

***Conclusions***:

As shown in the boxplot, the execution time decreases significantly as the number 
of processors increases, but beyond 8 cores, the performance gains plateaus. 
This suggests that using more than 8 cores offers little additional benefits.