---
title: "Building a background for PscanR"
author:
  - Federico Zambelli
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{5. Building a background for PscanR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
)

fixture <- function(...) {
    system.file(
        "extdata", "vignettes", "custom_background", ...,
        package = "PscanR", mustWork = TRUE
    )
}

run <- readRDS(fixture("ucbg_summary.rds"))
verification <- run$verification
fullbg <- run$fullbg

timing_of <- function(manifest, step) {
    manifest$elapsed_seconds[manifest$step == step]
}
timings_1core <- read.delim(
    fixture("timings_1core.tsv"), stringsAsFactors = FALSE
)
timings_8core <- read.delim(
    fixture("timings_8core.tsv"), stringsAsFactors = FALSE
)
build_speedup <- round(
    timing_of(timings_1core, "background_short") /
        timing_of(timings_8core, "background_short"), 1
)
full_bg_all_mb <- round(
    fullbg$size_bytes * verification$motif_count /
        length(fullbg$motifs) / 1024^2
)
is_cerevisiae <- vapply(
    TFBSTools::getMatrixSet(
        JASPAR2020::JASPAR2020,
        list(collection = "CORE", tax_group = "fungi")
    ),
    function(m) any(grepl("cerevisiae", TFBSTools::tags(m)$species)),
    logical(1)
)
n_cerevisiae <- sum(is_cerevisiae)
pct_cerevisiae <- round(100 * n_cerevisiae / length(is_cerevisiae))
```

# Overview

The PscanR package ships a collection of precomputed background sets for five model
organisms: human, mouse and fruit fly (RefSeq annotation), Arabidopsis (TAIR
annotation), and yeast (SGD annotation).

 `get_availableBG()` lists them:

```{r available}
head(get_availableBG(), 4)
```

This vignette describes how to build a custom background. Custom backgrounds 
are useful for cases when:

- **an organism is not in the catalogue**;
- **a promoter window is not in the catalogue**.

This vignette intentionally rebuilds one of the backgrounds that already exists 
in the shipped catalogue so that they *can* be compared for consistency: 
*Saccharomyces cerevisiae -950/+50*.

Every timing and verification figure quoted below was produced by `run_ucbg.R`,
which is kept in `UseCase/UC_Background/` of the PscanR source repository rather
than shipped with the package, because it requires network access and rebuilds
the whole background.

Yeast also keeps the demonstration short. On our server the whole build was
`r round(timing_of(timings_1core, "background_short"))` seconds on one core, for
`r format(verification$promoter_count, big.mark = ",")` promoters and
`r verification$motif_count` matrices; the same demonstration on human promoters
would be roughly 25 times that.

# Part A: build a background

## The promoters

A background is build from a set of sequences and a set of motif matrices. 
The sequences here are every promoter in the organism, defined exactly as the foreground promoters will later be defined: **−950 to +50 around the transcription start site**.

Three decisions are made here about the reference:

1. **the annotation**, UCSC's `sgdGene` table, here through the
   `TxDb.Scerevisiae.UCSC.sacCer3.sgdGene` package, whose version records
   *which* release was used;
2. **the genome**, `BSgenome.Scerevisiae.UCSC.sacCer3`, whose chromosome names
   must agree with the annotation's;
3. **the window**, 950 upstream and 50 downstream.

Of course, changing the reference will give different background measures, but 
usually slight updates in the gene annotation have minimal impact.

```{r promoters, eval=FALSE}

# UCSC's sgdGene table, packaged and versioned. 
txdb <- TxDb.Scerevisiae.UCSC.sacCer3.sgdGene::TxDb.Scerevisiae.UCSC.sacCer3.sgdGene

# The background covers the 16 nuclear chromosomes.
tx <- GenomicFeatures::transcripts(txdb, columns = c("tx_id", "tx_name"))
tx <- tx[as.character(GenomicRanges::seqnames(tx)) %in%
             paste0("chr", as.roman(1:16))]
names(tx) <- GenomicRanges::mcols(tx)$tx_name

# promoters() is strand-aware and anchored on the transcription start site.
ranges <- GenomicRanges::promoters(tx, upstream = 950, downstream = 50)

genome <- BSgenome.Scerevisiae.UCSC.sacCer3::BSgenome.Scerevisiae.UCSC.sacCer3

# A promoter running off the end of its chromosome cannot be read, and getSeq()
# would stop rather than skip it.
lengths <- GenomeInfoDb::seqlengths(genome)[
    as.character(GenomicRanges::seqnames(ranges))
]
inside <- GenomicRanges::start(ranges) >= 1L &
    GenomicRanges::end(ranges) <= lengths
ranges <- ranges[inside]

prom_seq <- Biostrings::getSeq(genome, ranges)
names(prom_seq) <- names(ranges)
```

That gives `r format(verification$transcript_count, big.mark = ",")`
promoters, that is one per transcript.

It is *not recommended* to drop duplicated (i.e. from transcripts sharing the same 
TSS), incomplete or partially sequenced promoters  since
the `ps_build_bg()` function will do the tidying itself,
dropping promoters that repeat one already in the set, promoters that are not
the full window, and promoters more than half unsequenced. 

The count that ends up in the file is what survived: the number of **distinct** promoters,
`r format(verification$promoter_count, big.mark = ",")` here.

## The matrices

Any `PFMatrixList` (see the TFBSTools package for details) will do. 
The shipped backgrounds use JASPAR CORE, restricted to the taxonomic group of the organism.

```{r matrices}
motifs <- TFBSTools::getMatrixSet(
    JASPAR2020::JASPAR2020,
    list(collection = "CORE", tax_group = "fungi")
)
length(motifs)
```

## Building a standard background

`ps_build_bg()` scans every sequence with every matrix, keeps the best-scoring
window per sequence per matrix, and reduces those to a mean and a standard
deviation.

```{r build, eval=FALSE}

background <- ps_build_bg(
    prom_seq, motifs,
    BPPARAM = BiocParallel::SerialParam()
)
```

```{r timing-table, echo=FALSE}
build_timings <- data.frame(
    Step = sprintf("ps_build_bg(), %d matrices", verification$motif_count),
    `One core` = timing_of(timings_1core, "background_short"),
    `Eight cores` = timing_of(timings_8core, "background_short"),
    check.names = FALSE
)
build_timings$Speedup <- round(
    build_timings$`One core` / build_timings$`Eight cores`, 1
)
knitr::kable(
    build_timings, digits = 1,
    caption = "Build time in seconds, MulticoreParam on eight cores."
)
```

Using eight cores (`BPPARAM` set to `BiocParallel::MulticoreParam(workers = 8)`)
was `r build_speedup`× faster on the full collection on our server. 
The gain scales with the size of the matrices collection, not with the promoter set.

## Writing it out, and reading it back

```{r write, eval=FALSE}
ps_write_bg_to_file(background, "J2020_sacCer3_950u_50d_UCSC.rebuild.txt")
```

This is the whole artefact:

```{r show-file}
bg_file <- fixture("J2020_sacCer3_950u_50d_UCSC.rebuild.txt")
cat(readLines(bg_file, n = 4), sep = "\n")
file.size(bg_file)
```

A header line, then one row per matrix: identifier, number of promoters,
mean score, standard deviation: standard PscanR backgrounds are small files.

`ps_retrieve_bg_from_file()` puts those numbers back onto the matrices they
belong to, and returns the `PSMatrixList` that `pscan()` expects.

```{r read-back}
background <- ps_retrieve_bg_from_file(bg_file, motifs)
background
head(ps_get_bg_table(background), 3)
```

The round trip closes here: `ps_get_bg_table()` returns exactly the three
columns that were written.

## The verification

The background just built should carry the same numbers as the one the package
ships for this combination, because it was built from the same annotation, the
same genome, the same window and the same matrices. We here show the result 
as obtained in the run on our server.


```{r verify-table}
knitr::kable(
    data.frame(
        Check = c(
            "Matrices compared",
            "BG_SIZE equal",
            "BG_MEAN agreeing to 10 decimal places",
            "BG_STDEV agreeing to 10 decimal places"
        ),
        Result = c(
            verification$motif_count,
            verification$size_equal,
            sprintf("%d / %d", verification$mean_agree_10dp,
                    verification$motif_count),
            sprintf("%d / %d", verification$stdev_agree_10dp,
                    verification$motif_count)
        )
    ),
    align = c("l", "r")
)
```

Every matrix agrees on both statistics. Following the recipe in this order gives the
catalogued numbers.

> Why ten decimal places and not the whole figure? How double precision numbers
> are handled differs between platforms (e.g. Linux x64 or Apple Silicon)
> leading to platform-dependent infinitesimal differences in the measured statistics.
> Those differences are well past where the enrichment test used by PscanR (z-test) 
> can notice and have no influence whatsoever on the enrichment results.

# Part B: building a full background (fullBG), and querying it by identifier

A standard background file keeps just three numbers per matrix, which means the
foreground has to be supplied as sequences: the promoters must be extracted
separately, with the right window.

PscanR also supports the creation and use of much bulkier full backgrounds
(fullBG). These allow a foreground to be scanned from identifiers alone rather
than from sequences, and they make foreground scanning very fast almost
independently of the number of promoters in the input. Their typical use case is
a large number of foreground scans against the same background.

A fullBG retains, for every promoter and every matrix, the hit score, position,
strand and matched oligonucleotide, and it builds a **legend** mapping every
original sequence name to the representative kept after de-duplication.

`pscan_fullBG()` then takes *transcript identifiers* as input where `pscan()`
takes sequences.

```{r build-full, eval=FALSE}

# The fifteen matrices this vignette keeps, and why: input/fullbg_motifs.tsv.
subset_ids <- read.delim(fixture("fullbg_motifs.tsv"))$MATRIX

full_bg <- ps_build_bg(
    prom_seq, motifs[subset_ids],
    BPPARAM = BiocParallel::SerialParam(), fullBG = TRUE
)
```


**Pass the sequences before de-duplication.** The legend is built from the names
of the sequences passed in, so passing the de-duplicated set produces a legend
that cannot resolve the names it dropped.

**Size.** A full background can generate a very large file (several gigabytes).  
This one is deliberately restricted to `r length(fullbg$motifs)` matrices, which brings it
to `r round(fullbg$size_bytes / 1024)` KB; scaling that by matrix count puts all
`r verification$motif_count` at roughly `r full_bg_all_mb` MB for yeast, and far
more for a vertebrate. For this reason, no FullBG is shipped with the package.

A full background is stored in a standard `.rds` file rather than a `.txt` one.

```{r load-full}
full_bg <- readRDS(fixture("sacCer3_950u_50d_fullBG_15motifs.rds"))
TFBSTools::name(full_bg)
```

## Example of scanning a foreground with a FullBG

Ribosomal protein genes are the textbook Rap1 target set in yeast.

The gene set is `r fullbg$rp_total` cytoplasmic ribosomal protein genes, taken
from `org.Sc.sgd.db` as standard names matching `^RP[LS][0-9]`, all
`r fullbg$rp_used` of them present in the background.

```{r foreground}
rp_promoters <- readRDS(fixture("rp_promoters.rds"))
rp_ids <- names(rp_promoters)
length(rp_ids)
```

```{r run-fullbg, message=TRUE}
by_legend <- pscan_fullBG(rp_ids, full_bg)
```

The message names the identifier convention PscanR detected in the background.

```{r fullbg-results}
enrichment <- ps_results_table(by_legend)
head(enrichment[, c("NAME", "ZSCORE", "P.VALUE", "FDR")], 5)
```

RAP1 comes first, at *Z* = `r round(fullbg$top_zscore, 1)`, an order of magnitude
above the next motif. SFP1, also reported on ribosomal protein promoters, is a
distant second.


> What a FullBG buys is time and interface.
> `pscan_fullBG()` reads hits that were computed once when the background was
> built, so asking about a gene set costs a lookup in the legend and a subset of
> the stored scores, whatever the size of the foreground. 
> The input of a FullBG foreground scan is a list of gene identifiers, instead 
> of having to reproduce the promoter extraction.


# Part C: a checklist for building a new background

In order:

1. **Annotation source.** A packaged `TxDb` if Bioconductor ships one for the
   organism. Otherwise a UCSC table via `txdbmaker::makeTxDbFromUCSC()`, or a
   GFF/GTF via `txdbmaker::makeTxDbFromGFF()`; with those, record *which
   release* was used and when it was retrieved. This is the single decision that
   determines whether the background is reproducible.
2. **Genome sequence.**
   A `BSgenome` package if one exists; otherwise a FASTA read with
   `Biostrings::readDNAStringSet()`. Make sure the chromosome names convention
   used in the genome is the same as that in the annotation.
3. **Window.** A free parameter, and the one thing that must be identical between
   background and foreground. 
4. **Motif collection, and its taxonomic subset.** Inspect what the collection
   actually contains for the organism. For example, JASPAR CORE `fungi`, used
   above, is `r pct_cerevisiae`% *S. cerevisiae*, so it may or may not be
   suitable for other fungi.
5. **Identifier scheme.** Which suffixes, if any, are part of the identifier.
   `pscan_fullBG(scheme = )` names the grammar explicitly; `"auto"` detects it and
   reports what it found but currently it supports only RefSeq, SGD and TAIR 
   name conventions.

# Summary

```{r workflow, eval=FALSE}

# 1. Promoters, in the window the foreground will use

genome <- BSgenome.Scerevisiae.UCSC.sacCer3::BSgenome.Scerevisiae.UCSC.sacCer3

txdb <- TxDb.Scerevisiae.UCSC.sacCer3.sgdGene::TxDb.Scerevisiae.UCSC.sacCer3.sgdGene

tx <- GenomicFeatures::transcripts(txdb, columns = c("tx_id", "tx_name"))
tx <- tx[as.character(GenomicRanges::seqnames(tx)) %in%
             paste0("chr", as.roman(1:16))]
names(tx) <- GenomicRanges::mcols(tx)$tx_name

ranges <- GenomicRanges::promoters(tx, upstream = 950, downstream = 50)

lengths <- GenomeInfoDb::seqlengths(genome)[
    as.character(GenomicRanges::seqnames(ranges))
]

inside <- GenomicRanges::start(ranges) >= 1L &
    GenomicRanges::end(ranges) <= lengths

ranges <- ranges[inside]

prom_seq <- Biostrings::getSeq(genome, ranges)

# 2. Matrices, in the taxonomic group of the organism.
motifs <- TFBSTools::getMatrixSet(
    JASPAR2020::JASPAR2020, list(collection = "CORE", tax_group = "fungi")
)

# 3. Scan.
background <- ps_build_bg(
    prom_seq, motifs, BPPARAM = BiocParallel::SerialParam()
)

# 4. Keep it.
ps_write_bg_to_file(background, "my_background.txt")
background <- ps_retrieve_bg_from_file("my_background.txt", motifs)

# Optionally, keep the scan as well, and query it by identifier.
full_bg <- ps_build_bg(prom_seq, motifs, fullBG = TRUE)

# The identifiers of the genes of interest, in the scheme the background
# uses.
my_gene_ids <- readLines("my_gene_ids.txt")

ps_results_table(pscan_fullBG(my_gene_ids, full_bg))
```

# Session information

```{r session-info}
sessionInfo()
```
