---
title: "PscanR Concepts and Input Preparation"
author:
    - Diana Betelli
    - Federico Zambelli
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
vignette: >
    %\VignetteIndexEntry{2. PscanR Concepts and Input Preparation}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    markdown:
        wrap: 80
---

```{r setup, include = FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Introduction

**PscanR** is an R implementation of the algorithm developed by Zambelli et al.
(2009) for analyzing sequence motif enrichment. It helps identify transcription
factors (TFs) that are potential regulators of a set of co-regulated or
co-expressed genes. By scanning input sequences (e.g., promoters) against a
database of known TF binding motifs, **PscanR** assesses which motifs are
significantly over- or under- represented and highlights candidate binding
sites.

## Workflow overview:

***Input***: A set of DNA sequences (e.g., promoter regions) and a library of
known transcription factors binding motif, such as Position Weight Matrices
(PWMs) from the JASPAR database.

***Analysis***: Identification of enriched motif and calculation of statistical
significance (Z-scores and P-Values).

***Output***: A ranked list of enriched motif, along with their associated
P-Values and Z-scores.

# The Background Model

In order to assess the significance of the obtained result, a background model
must be computed in the form of a PSMatrixList object. 

In PscanR the input sequences are considered a sample drawn from a universe 
consisting of all promoter sequences available for the species under
investigation. This *universe* serves as the background model for statistical
analyses, ensuring a comprehensive and unbiased assessment of the analysis
outcomes.

Because the background is the distribution of scores over the whole promoter
universe, the foreground and the background must describe the *same* kind of
sequence. A promoter window of 200 bp upstream and 50 bp downstream of the TSS
does not yield scores comparable with one of 950 bp upstream and 50 bp
downstream: a longer window offers a motif more places to match, so its best
score rises even when nothing is enriched. The foreground promoters must
therefore be extracted with exactly the window the background was built on, and
each shipped background records that window in its name, as in `950u_50d`.

The next section shows how to use one of the pre-computed backgrounds available
for PscanR. Please refer to the the vignette *Building a background for PscanR* 
for a walkthrough on how to build and use a custom background.

## Obtaining a Pre-Computed Background

Most analyses do not need a background to be built, because one has been
computed already. `get_availableBG()` lists what is available, and
`generate_psmatrixlist_from_background()` retrieves one and returns it as the
`PSMatrixList` that `pscan()` takes:

```{r list-backgrounds, eval=FALSE}
get_availableBG()
get_availableBG(details = TRUE) # the full version-2 catalog
```

```{r download-background, eval=FALSE}
J2020_PSBG <- generate_psmatrixlist_from_background(
    "Jaspar2020", "hs", c(-200, 50), "hg38"
)
```

Four things identify a background, and they are the four arguments above: the
JASPAR release the motifs come from, the species, the promoter window relative
to the TSS, and the genome assembly. The assembly is required only for *Mus
musculus* and *Homo sapiens*, which have more than one. Available combinations
are:

-   *Homo sapiens*, assemblies hg38 and hs1

-   *Mus musculus*, assemblies mm10 and mm39

-   *Arabidopsis thaliana*, TAIR9

-   *Drosophila melanogaster*, dm6

-   *Saccharomyces cerevisiae*, sacCer3

with promoter windows of 200 bp upstream and 50 bp downstream, 450 upstream and
50 downstream, 500 upstream and 0 downstream, 950 upstream and 50 downstream,
and 1000 upstream and 0 downstream, for the JASPAR 2020, 2022 and 2024
releases. Only JASPAR **core** motifs are used.

The call above needs no argument naming where the data comes from, because the
retrieval is arranged so that it does not have to. It resolves the resource
through ExperimentHub, and falls back to the immutable Zenodo record
<https://doi.org/10.5281/zenodo.21821764> with a warning if the Hub cannot be
reached, so a temporary Hub outage does not stop an analysis. Both routes serve
the same archive and every file is verified by SHA-256 on arrival. Pass
`source = "zenodo"` to go to Zenodo directly, or `source = "github"` to reach
the legacy version-1 resources, which no other backend exposes.

For a self-contained example, the package ships the JASPAR2020 core matrices
and the matching human 200-up, 50-down background as a table, which
`ps_retrieve_bg_from_file()` combines into the same kind of object without any
download:

```{r offline-background}
J2020 <- readRDS(system.file("extdata", "J2020.rds", package = "PscanR"))
bg_path <- system.file(
    "extdata", "J2020_hg38_200u_50d_UCSC.psbg.txt",
    package = "PscanR"
)
J2020_PSBG <- ps_retrieve_bg_from_file(bg_path, J2020)
J2020_PSBG[[4]] # one matrix, with its background statistics
```

This is the background the worked example at the end of this vignette scans
against. When no shipped background matches the species, window or motif
collection an analysis needs, one can be built instead; the vignette *Building
a background for PscanR* covers that case.

# The Input Transcript For `pscan()`

**`pscan()`** accepts as input a DNAStringSet object containing the DNA
sequences of co-regulated or co-expressed genes. This object can be built from a
list of gene/transcript identifiers.

For humans, mice, and Drosophila, the available background models are based on
RefSeq identifiers, for Arabidopsis on TAIR, and for yeast on SGD IDs. Other
types of gene identifiers can also be used if the corresponding background model
is created; the *Building a background for PscanR* vignette covers that case.

Below is an example dataset containing **human NRF1 target transcripts**
(RefSeq), curated from the literature. Since PscanR works at the transcript
level, gene identifiers must be converted to their corresponding transcript IDs
before analysis. The number **100** indicates that 100% transcripts in the list
correspond to known NRF1 target.

```{r chunk-2}
file_path <- system.file("extdata", "nrf1100.txt", 
                            package = "PscanR")
target <- read.csv(file_path, header = FALSE)
head(target)
```

## From Transcript IDs to promoter sequences

PscanR analysis works at the promoter level, and Transcript IDs provide a 
univoque match for the corresponding promoter within the same assembly and gene 
annotation system. 

However, since it is more common to obtain from standard bioinformatic workflows
lists of co-regulated or co-expressed genes rather than transcripts , we provide 
a way  to help users in the reproducible conversion from Gene IDs to Transcript 
IDs for use with PscanR.

`ps_select_promoters()` performs that task.

The promoters shipped with the package are human, 200 bp upstream and 50 bp
downstream of the TSS, which is the window of the background loaded earlier:

```{r load-promoters}
library(Biostrings)
prom_seq <- readRDS(
    system.file("extdata", "prom_seq.rds", package = "PscanR")
)
length(prom_seq)
unique(width(prom_seq)) # 200 upstream + 50 downstream
head(names(prom_seq), 3)
```

`ps_select_promoters()` is given the genes of interest, the promoters to choose
from, and a table relating one to the other. The NRF1 targets read above are
RefSeq transcripts, so `org.Hs.eg.db` supplies the gene symbol behind each of
them and the table is built from the result:

```{r build-annotation}
library(org.Hs.eg.db)
annotation <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = target[[1]], keytype = "REFSEQ", columns = "SYMBOL"
)
annotation <- annotation[!is.na(annotation$SYMBOL), ]
genes <- unique(annotation$SYMBOL)
length(genes)
```

Note that the identifiers in this table carry no version, whereas the promoters
are named `NM_006559.3` and do. This mismatch is the reason the function
inspects the identifiers before matching them rather than comparing them as
strings. It recognises the RefSeq convention, in which the trailing number is a
release version and two versions name the same transcript, and normalises the
suffix away. It does not do so for TAIR identifiers such as `AT1G01110.2`,
where the same punctuation marks a splice variant that is a different
transcript, frequently with a different start site.

```{r select-promoters}
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    mode = "representative",
    return = "both"
)
```

The mapping records how each promoter was arrived at, which keeps the choice
auditable rather than implicit:

```{r selection-mapping}
head(
    selection$mapping[, c(
        "gene", "transcript_id", "promoter_id",
        "scheme", "suffix_role", "n_candidates", "decided_by"
    )],
    3
)
```

`ps_selection_summary()` reports the same thing over the whole set, including
how many genes offered a real choice and how many were settled by the tie-break
rather than by the ranking:

```{r selection-summary}
ps_selection_summary(selection$mapping)
```

One gene is unmapped because its transcript is absent from this small promoter
set. The remaining 90 are returned as a `DNAStringSet`, named by gene and
transcript so that a result can be traced back to both:

```{r selected-sequences}
foreground <- selection$sequences
foreground
```

`mode = "representative"` ranks transcripts by the scheme's own rule, which for
RefSeq prefers `NM_` over `NR_` over the predicted `XM_` and `XR_` accessions.
For human hg38 a better ranking is available, because MANE Select and RefSeq
Select name one transcript per gene by curation. `mode = "select"` uses it:

```{r mane-select, eval=FALSE}
select_tx <- ps_load_select_transcripts("hg38")
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    select_transcripts = select_tx,
    mode = "select",
    return = "both"
)
```

That chunk is not run when this vignette is built, because
`ps_load_select_transcripts()` downloads the table from UCSC and caches it. The
metadata it retrieves is specific to human hg38, so for other organisms either
`mode = "representative"` applies, or a species-specific table is supplied
through `select_transcripts`.

The foreground is now what `pscan()` takes, together with the background from
the earlier section:

```{r ready-for-pscan, eval=FALSE}
res <- pscan(foreground, J2020_PSBG)
```

Both arguments describe the same 200-up, 50-down window, which is what makes
their scores comparable. Had the background been a `950u_50d` one, these
promoters would have had to be extracted again over 950 bp upstream rather than
200, because a longer window gives a motif more places to match and raises its
best score whether or not anything is enriched.

# Choosing a `BPPARAM`

`pscan()` parallelises over motifs through **BiocParallel**, and the `BPPARAM`
argument selects the strategy. The recommendation is:

-   **`MulticoreParam(n)` on Linux and macOS.** It parallelises by forking, which
    is the faster of the two because the workers inherit the session's memory
    instead of receiving a copy of it.

-   **`SnowParam(n)` where forking is unavailable**, which in practice means
    Windows. It starts independent processes and communicates over sockets, so
    it works everywhere at the cost of the data transfer.

-   **`SerialParam()` when debugging**,   since it starts no workers at all and a
    failure therefore surfaces in the session rather than inside one of them.

The default is `bpparam()`, which picks a strategy and a worker count from the
session's own configuration.

Raise the number of cores when a scan on one appears slow. More cores may not
help on a scan that is already quick, because starting the workers and shipping
the matrices out to them costs time that a short scan does not repay.

# Session information

```{r session-info}
sessionInfo()
```
