---
title: "Motif enrichment analysis with PscanR: Arabidopsis BPC target promoters"
author:
  - Federico Zambelli
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{4. Motif enrichment analysis with PscanR: Arabidopsis BPC target promoters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
)

fixture <- function(...) {
    system.file(
        "extdata", "vignettes", "arabidopsis_bpc", ...,
        package = "PscanR", mustWork = TRUE
    )
}
```

# Overview

PscanR looks for transcription factor binding motifs that are over-represented
in the regulatory regions of a set of genes. The idea in brief: take the
promoters of genes that behave alike (co-expressed, co-regulated, differentially
expressed in the same contrast), scan them with a collection of position
frequency matrices, and ask which motifs score higher there than they do across
all promoters of the organism, that is the universe background set. 

The package ships a collection of precomputed background sets for five model
organisms: human, mouse and fruit fly (RefSeq annotation), *Arabidopsis* (TAIR
annotation), and yeast (SGD annotation). The package also provides functions to
compute and use custom backgrounds, for those or for different organisms (not
covered in this vignette).

This vignette works through a PscanR analysis on A. thaliana promoters
and through several functions the package provides for inspecting the result.

## The data

The example uses gene expression data from a study of BASIC PENTACYSTEINE (BPC)
transcription factors, which control differentiation of the medial domain of
the *Arabidopsis* gynoecium.

> Caselli F, Palermiti M, Petrella R, Morlacchi VA, Dünser K, Kleine-Vehn J,
> Chiara M, Gregis V. *A novel role of BPCs in the control of medial domain
> differentiation during gynoecium development in Arabidopsis thaliana.*
> Plant Mol Biol 116(1):4 (2025). doi:10.1007/s11103-025-01662-x

The input is a two-column table holding a gene identifier and the direction of
change, which is all an analysis needs to begin.

Every fixture this vignette reads was produced by `run_uc4_pscan.R`, which is
kept in `UseCase/UC4_Arabidopsis_TransmittingTract/` of the PscanR source
repository rather than shipped with the package, because it requires network
access and the full TAIR9 annotation.

```{r read-gene-sets}
gene_sets <- read.delim(
    fixture("arabidopsis_bpc_gene_sets.tsv"),
    stringsAsFactors = FALSE
)
str(gene_sets)
table(gene_sets$direction)
```

The rest of this vignette follows the `r sum(gene_sets$direction == "down")`
downregulated genes, those that lose expression when BPC function is impaired.

# The background

PscanR precomputed backgrounds are indexed by organism, genome assembly, promoter window and motif
collection. `get_availableBG()` lists what is available.

```{r discover-background, eval=FALSE}
get_availableBG(keyword = "TAIR9", details = TRUE)
```

`generate_psmatrixlist_from_background()` retrieves a background and attaches it to the
matching JASPAR collection, returning a `PSMatrixList`, that is the object PscanR works
with throughout.

```{r retrieve-background, eval=FALSE}
background <- generate_psmatrixlist_from_background(
    JASPAR_matrix = "Jaspar2020",
    org = "at",
    prom_reg = c(-1000, 0),
    assembly = "TAIR9",
    version = 2
)
length(background)
#> [1] 530
```

That returns the JASPAR 2020 CORE *plants* collection, 530 matrices, each
carrying three numbers: how many promoters its background was computed
over, and the mean and standard deviation of its score across them.

Take notice, the promoter window selected, 1000 bp upstream of the transcription start
site here, is important: **to be comparable, the foreground promoters must be extracted
with the same promoter window the background was built from**. 

The requirement is that foreground and background *match each other*. 

# From genes to promoters

The `pscan()` function of PscanR takes as input a `DNAStringSet` object containing a set of promoters from  co-regulated or co-expressed genes. The user needs to take care to match the promoter
window of the selected background. 

> The ideal workflow would require knowing which transcripts are co-regulated or
> co-expressed, in order to retrieve the exact set of promoters regulating their
> expression. Usually, though, the output of a differential expression analysis
> is a list of genes rather than transcripts.
>
> Since a gene may have several transcription start sites and therefore
> different promoters, when working at the gene level the choice of which
> promoter to use for each gene is necessarily arbitrary. This is not ideal, but
> it removes the barrier of having to work at the transcript level, which
> remains the preferred option.

The `ps_select_promoters()` function is a utility function that can be used to pass from gene or 
transcript identifiers to their promoter sequences for the supported organisms and annotations,  deterministically resolving a gene list to one promoter per gene.

Before it can do that it has to work out what the identifiers mean, because the
same punctuation carries different meanings in different references. In RefSeq,
`NM_000546.6` ends in a release **version**, so that two versions name one
transcript. In TAIR, `AT1G01110.2` ends in a **splice variant**, which is a
different transcript,
often with a different start site. PscanR detects the convention and reports it.

Toy example:

```{r promoter-scheme, message=TRUE}
promoters <- c("AT1G01110.2", "AT1G01110.1", "AT1G01160.1")
annotation <- data.frame(
    GENE = sub("\\..*$", "", promoters),
    TX = promoters
)

selection <- ps_select_promoters(
    genes = c("AT1G01110", "AT1G01160"),
    promoter_ids = promoters,
    annotation = annotation,
    gene_col = "GENE", transcript_col = "TX",
    mode = "representative"
)
```

The returned mapping records the reason for the choice:

```{r promoter-scheme-table, echo=FALSE}
knitr::kable(
    selection[, c(
        "gene", "promoter_id", "scheme", "suffix_role",
        "transcript_class", "selection_source", "n_candidates", "decided_by"
    )],
    caption = "One promoter per gene, and why each was chosen."
)
```

For *Arabidopsis* the rule is to prefer the `.1` primary gene model and fall
back to the lowest available variant.

For our dataset we need promoter sequences matching the background selected
above: TAIR9 annotation, 1000 bp upstream of the TSS.

Notice Bioconductor's `TxDb.Athaliana.BioMart.*` packages are
Ensembl Plants builds of the newer Araport11 gene models, which extended many
5' UTRs of the transcript identifiers they share with TAIR9. Those extensions
*would* shift transcription start sites, and with them the promoters. 
Since the background is TAIR9, so is the annotation used here.

The following chunk shows how to do it. It is not evaluated during the vignette
build, because it requires network access, and it additionally needs the
`txdbmaker`, `GenomicFeatures`, `GenomicRanges`, `GenomeInfoDb`, `Biostrings`
and `BSgenome.Athaliana.TAIR.TAIR9` packages.

`TAIR9_GFF3_genes.gff` can be downloaded from the
[TAIR9 GFF3 release directory](https://www.arabidopsis.org/download/list?dir=Genes/TAIR9_genome_release/tair9_gff3).
Set `tair9_gff` to the path it was saved to.

```{r select-real, eval=FALSE}

down_genes <- gene_sets$gene_id[gene_sets$direction == "down"]

# The same TAIR9 annotation the background was built from.

tair9_gff <- "TAIR9_GFF3_genes.gff" # path to the downloaded annotation

txdb <- txdbmaker::makeTxDbFromGFF(
    tair9_gff, format = "gff3",
    dataSource = "TAIR9", organism = "Arabidopsis thaliana"
)
genome <- BSgenome.Athaliana.TAIR.TAIR9::BSgenome.Athaliana.TAIR.TAIR9

# The background covers the five nuclear chromosomes.
tx <- GenomicFeatures::transcripts(txdb, columns = "tx_name")
tx <- tx[as.character(GenomicRanges::seqnames(tx)) %in% paste0("Chr", 1:5)]
names(tx) <- GenomicRanges::mcols(tx)$tx_name

# promoters() takes the region upstream of each 5' end, so it is strand-aware
# and anchored on the transcription start site.
ranges <- GenomicRanges::promoters(tx, upstream = 1000, downstream = 0)

# A promoter that runs off the end of its chromosome cannot be read, and
# getSeq() would stop rather than skip it. One transcript is affected here.
inside <- GenomicRanges::start(ranges) >= 1L &
    GenomicRanges::end(ranges) <= GenomeInfoDb::seqlengths(genome)[
        as.character(GenomicRanges::seqnames(ranges))
    ]
ranges <- ranges[inside]

promoters_tair9 <- Biostrings::getSeq(genome, ranges)
names(promoters_tair9) <- names(ranges)

# One more promoter is more than half unsequenced.
promoters_tair9 <- promoters_tair9[
    Biostrings::letterFrequency(promoters_tair9, "N", as.prob = TRUE) <= 0.5
]

annotation <- data.frame(
    GENE = sub("\\..*$", "", names(promoters_tair9)),
    TX = names(promoters_tair9)
)

down_mapping <- ps_select_promoters(
    genes = down_genes,
    promoter_sequences = promoters_tair9,
    annotation = annotation,
    gene_col = "GENE", transcript_col = "TX",
    mode = "representative",
    return = "mapping"
)

down_promoters <- promoters_tair9[down_mapping$promoter_id]
```

That mapping was saved along with the scan it fed, so the rest of this vignette
can read it directly from the precomputed object.

```{r load-precomputed}
precomputed <- readRDS(fixture("arabidopsis_bpc_results.rds"))
down_mapping <- precomputed$down_mapping
```

`ps_selection_summary()` condenses the gene-to-transcript mapping into something reviewable:
how many genes were asked for, how many were placed, and how often there was a
real choice to make.

```{r selection-summary}
ps_selection_summary(down_mapping)$overall
```

The genes that go unmapped are usually loci annotated after the assembly was released.
`selection_source` names the rule that decided each promoter:

```{r selection-by-source}
ps_selection_summary(down_mapping)$by_source
```

# Running the scan

`pscan()` takes the foreground promoter sequences and the background `PSMatrixList`, and
returns the updated `PSMatrixList` carrying the per-sequence results for every motif.

The BPPARAM from `BiocParallel` is useful to set the level of parallelism:

- `SerialParam()` uses one R process and is the safest choice for package checks and reproducible debugging. 
 
- `SnowParam()` starts independent socket-based R workers; it is the portable parallel choice on Windows and Unix-like systems, but has greater worker startup and memory overhead. 
 
- `MulticoreParam()` uses lower-overhead forked workers and is generally preferable for local parallel execution on Linux or other supported Unix-like systems; it should not be used on Windows.

The next chunk is not evaluated by default to limit the vignette's build time, but the user can
run it. It should take under two minutes with `BPPARAM` set to
`BiocParallel::SerialParam()`, or less with parallelism enabled.

```{r scan, eval=FALSE}

BPPARAM <- BiocParallel::SerialParam() # Serial - Compatible but slower

# Cross-platform parallel execution with four independent R workers:
# BPPARAM <- BiocParallel::SnowParam(workers = 4, type = "SOCK") #Use this on Windows

# Lower-overhead forked execution on supported Unix-like systems: #Use this on Linux and MacOS
# BPPARAM <- BiocParallel::MulticoreParam(workers = 4) 

results <- pscan(
    down_promoters, background,
    BPPARAM = BPPARAM
)
```

`ps_results_table()` turns that object into the enrichment table.

```{r results-table, eval=FALSE}
enrichment <- ps_results_table(results)
```


To shorten the vignette's build time the scan was saved into the same RDS file the mapping came from. The two objects below are that saved result; everything from here on is computed from them.

```{r load-results}
# Exactly what ps_results_table() returned for the downregulated promoters.
enrichment <- precomputed$enrichment_down
dim(enrichment)

# The same scan kept as a PSMatrixList rather than a table, so the
# per-promoter detail survives, restricted to a few motifs to keep it small.
results_subset <- precomputed$results_subset
results_subset
```

Those motifs were chosen to span the biology of the original study: the BPC factors it
is about, the G-box binders that turn out to dominate the result, and two unrelated motifs for contrast.

```{r subset-names}
TFBSTools::name(results_subset)
```

The enrichment table is one row per motif, keyed by JASPAR matrix identifier:

```{r results-table-shown}
head(enrichment[order(-enrichment$ZSCORE), ], 6)
```

The top of the full table is occupied by G-box binders of the bHLH and bZIP
families: ABI5, ABF1 and ABF2 are all bZIPs recognising the ACGT core.

The background parameters travel with the result, so they can be read back
from it at any point with `ps_get_bg_table()`:

```{r bg-table}
ps_get_bg_table(results_subset)
```

`BG_SIZE` is the number of *Arabidopsis* promoters the background was computed
over; `BG_MEAN` and `BG_STDEV` describe how that motif scores across them. 

Each row of the enrichment table compares the foreground mean score against the
background mean, turns the difference into a z-score, and reports the corresponding p-value with a
Benjamini–Hochberg FDR across the collection. 

`ps_motif_barplot()` ranks the motifs and draws them; the `n` argument limits the
plot to the top of the list (default to at most 20). `group = "class"` colours each bar by the
structural class of the factor, which `ps_motif_class()` reads off the JASPAR matrices.

```{r top-motifs, fig.width=8, fig.height=5, dpi=120}

table(ps_motif_class(results_subset))

ps_motif_barplot(results_subset, group = "class")
```

Within the subset the G-box binders rank above the three BPC motifs, as they do
in the full table above.

The function returns a `ggplot` object rather than drawing immediately, so a
title, a different theme or any other layer can be added afterwards:

```{r barplot-composable, eval=FALSE}
ps_motif_barplot(results_subset, group = "class") +
    ggplot2::labs(title = "Downregulated promoters")
```

# What a result object contains

`ps_results_table()` summarises across promoters, but the `PSMatrixList` keeps
the per-promoter detail and can be explored with the accessors.

A `PSMatrixList` is indexed by JASPAR matrix identifier, so `MA0931.1` pulls
out the entry for ABI5:

```{r accessors}
abi5 <- results_subset[["MA0931.1"]]
TFBSTools::name(abi5)

ps_zscore(abi5)
ps_pvalue(abi5)
c(background = ps_bg_avg(abi5), foreground = ps_fg_avg(abi5))
ps_hits_size(abi5)
```

`ps_hits_table()` returns the best hit in every promoter: its score, position,
strand, and the matched oligonucleotide.

```{r hits-table}
head(ps_hits_table(abi5), 5)
```

For example, for ABI5 the recovered sequences are variations on the G-box core `CACGTG`.

# Structure across promoters

`ps_z_table()` gives the per-promoter z-scores as a matrix.

```{r z-table}
z <- ps_z_table(results_subset)
dim(z)
round(z[1:4, 1:5], 2)
```

`ps_zscore_heatmap()` clusters that matrix, which exposes how motifs group
across the promoter set.

```{r zscore-heatmap, fig.width=9, fig.height=7, dpi=110}
ps_zscore_heatmap(
    results_subset, FDR = 0.01,
    show_rownames = FALSE, treeheight_row = 0
)
```

In this case, the column dendrogram roughly splits the motifs in two. The three BPC paralogues sit
together, unsurprisingly since they recognise very similar GA-rich elements.
Opposite them is a block of G-box binders (HY5, the ABF/AREB bZIPs, the PIF and
SPT bHLHs). Promoters scoring high across the BPC columns
are largely not the ones scoring high across the G-box columns, and vice versa,
something that could not be seen looking only at the enrichment table.

# Where the sites are

 `ps_hitpos_map()` plots the position of the best hit in each promoter.

```{r hitpos-map, fig.width=9, fig.height=7, dpi=110}
ps_hitpos_map(
    results_subset, FDR = 0.01, shift = -1000,
    show_rownames = FALSE, treeheight_row = 0
)
```

`shift = -1000` relabels the axis so that coordinates are relative to the
transcription start site, matching how the promoters were extracted.

For a single motif, `ps_density_plot()` shows the distribution of hit positions
directly.

```{r density-plot, fig.width=7.5, fig.height=4.6, dpi=120}
ps_density_plot(
    abi5,
    shift = -1000, st = "strict",
    window = c(-1000, 0), bins = 20
)
```

Strong ABI5 sites are concentrated just upstream of the transcription start
site, peaking within about 100 bp of it, rather than spread evenly over the
kilobase scanned.

`window` defines the promoter window the positions came from. 
Given the window, the kernel density estimate is corrected by reflection. The
reflecting bounds are the positions a hit can be *reported* at: a hit is
recorded at the start of the motif, so it can never fall in the last few bases
of the window.

`bins` adds the same positions as a binned profile. Bars are on the density
scale, so they share the curve's axis directly. 

The profile is worth comparing between motifs. BPC6 binds GA-rich elements rather than the ACGT core ABI5 recognises.

```{r density-plot-bpc6, fig.width=7.5, fig.height=4.6, dpi=120}
bpc6_hits <- results_subset[["MA1402.1"]]

ps_density_plot(
    bpc6_hits,
    shift = -1000, st = "strict",
    window = c(-1000, 0), bins = 20
)
```

Both motifs are concentrated near the transcription start site, but not in the
same way. ABI5 peaks around 75–100 bp upstream and then falls away over the last
stretch; BPC6 climbs monotonically and is densest in the final bin, right up
against the start site. Note where its curve stops: the motif is
`r ncol(bpc6_hits)` bp wide and a hit is reported at its start, so it cannot be
placed closer than `r ncol(bpc6_hits) - 1` bp from the end of the window. 

The `st` argument sets which hits are used: `"all"` uses every promoter, `"loose"`
those scoring above the motif's background mean, and `"strict"` those a standard
deviation above it. Restricting to strong sites is usually what makes positional
structure visible, because with every promoter included the weak matches present in
almost any sequence flatten the distribution.

`ps_hit_score_plot()` puts one point per promoter at its hit
position and its hit score, with the two thresholds `st` names drawn as references.

```{r hit-score-plot, fig.width=9.5, fig.height=4.6, dpi=115}
ps_hit_score_plot(
    results_subset[c("MA1402.1", "MA0931.1")],
    shift = -1000
)
```

The two motifs behave differently. BPC6's retained hits bank into the last few
hundred bases while its weak hits stay spread across the kilobase, so its
positional concentration is specifically a concentration of *strong* sites.
ABI5's retained hits are spread more evenly, and sit somewhat further from the
transcription start site than BPC6's.

# Narrowing the question with a filter

The scan so far asked what is enriched across the whole set of downregulated promoters.
A narrower question is what is enriched *in the subset of those promoters that carry a
strong match for a particular factor*.

`pscan_filtered()` scans the promoters with one chosen
matrix, keeps those scoring at least `n` standard deviations above that motif's
background mean, and rescans the survivors against the whole collection.

The following chunk is not evaluated, to limit the vignette's build time, but it
can be run as it stands and should take under a minute.

```{r pscan-filtered, eval=FALSE}
bpc6 <- background[[which(TFBSTools::ID(background) == "MA1402.1")]]

filtered <- pscan_filtered(
    down_promoters, bpc6, n = 1,
    background = background,
    BPPARAM = BiocParallel::SerialParam()
)
ps_results_table(filtered)
```

Anchoring on BPC6 keeps `r precomputed$filtered_bpc6$n_retained` of the
`r precomputed$filtered_bpc6$n_input` downregulated promoters.

```{r filtered-table, echo=FALSE}
ft <- precomputed$filtered_bpc6$table
ft <- head(ft[order(-ft$ZSCORE), ], 10)
knitr::kable(
    data.frame(
        motif_id = ft$motif_id, NAME = ft$NAME,
        ZSCORE = round(ft$ZSCORE, 2),
        P.VALUE = format(ft$P.VALUE, scientific = TRUE, digits = 3)
    ),
    row.names = FALSE, align = c("l", "l", "r", "r"),
    caption = paste(
        "Enrichment within the", precomputed$filtered_bpc6$n_retained,
        "promoters carrying a strong BPC6 site."
    )
)
```

The BPC motifs and RAMOSA1, all of them GA-repeat binders, now sit far above
everything else, which is the expected consequence of selecting on one of them.
What is informative is the rest of the list: the G-box bZIP and bHLH motifs that
dominated the unfiltered scan have dropped out of the top entirely, replaced by
C2H2 and Dof factors. Apparently, the promoters BPC is positioned to bind are a different
population from the ones driving the overall G-box signal.

That is the point of the filter: it separates a mixed foreground into
subpopulations that a single scan reports only in aggregate.

# Summary

A complete PscanR analysis is four steps: choose a background, resolve genes to
promoters with a matching window, scan, and read the table.

```{r workflow, eval=FALSE}
background <- generate_psmatrixlist_from_background(
    "Jaspar2020", "at", c(-1000, 0), "TAIR9", version = 2
)
down_promoters <- ps_select_promoters(
    genes = down_genes, promoter_sequences = promoters_tair9,
    annotation = annotation, gene_col = "GENE", transcript_col = "TX",
    mode = "representative", return = "sequences"
)
results <- pscan(down_promoters, background)
ps_results_table(results)
```

Everything after that is interpretation: per-promoter scores (`ps_z_table()`),
hit positions and sequences (`ps_hits_table()`, `ps_hitpos_map()`,
`ps_density_plot()`, `ps_hit_score_plot()`), and the ability to restrict the
foreground to promoters carrying a particular site (`pscan_filtered()`).

# Appendix: why this analysis agrees closely, but not exactly, with the published one

The study used here reports its own Pscan results, so the two can be compared
directly. They agree on the biology and on most of the motif list: of the 168
motifs called significant in the published downregulated set, 164 are recovered
here. Three differences between the setups account for the rest, none of them a
disagreement about the data.

**The promoter window.** The published analysis used −1000/+100 relative to the
TSS. No precomputed background currently exists for that window, so this one
uses −1000/0: the same upstream extent, without the 100 bp downstream. That
100 bp is unlikely to matter much, judging by a comparison that can be made:
scanning the same genes against the −950/+50 background instead yields z-scores
correlating with these at ρ = 0.99 across all 530 motifs. Shifting the window by
50 bp barely moves the underlying statistic, even though it does move some calls
across the significance threshold.

**The annotation release.** The published promoters were defined on TAIR10 gene
models; the background here was built from TAIR9. This is smaller than it
sounds, because TAIR10 is an annotation release on the *same genome assembly*:
it adds loci and gene models without moving coordinates. The consequence is not
shifted promoters but a few genes that do not yet exist in the older
annotation: 2244 of the study's 2253 genes are present in the TAIR9 promoter
universe, or 99.6 %. The nine missing are mostly ncRNA loci introduced in
TAIR10, plus one mitochondrial gene outside the nuclear promoter set, and they
are split between the up and down lists rather than concentrated in one.

**Multiple testing correction.** PscanR reports the raw *P*-value alongside a
Benjamini–Hochberg FDR. The published significance values follow a different
convention, so the two lists were put on the same footing before being compared.
Since the raw *P*-value travels with the result table, any correction can be
applied to it afterwards.

# Session information

```{r session-info}
sessionInfo()
```
