---
title: "PscanR Concepts and Input Preparation"
author:
    - Diana Betelli
    - Federico Zambelli
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
vignette: >
    %\VignetteIndexEntry{2. PscanR Concepts and Input Preparation}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    markdown:
        wrap: 80
---

```{r setup, include = FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Introduction

**PscanR** is an R implementation of the algorithm developed by Zambelli et al.
(2009) for analyzing sequence motif enrichment. It helps identify transcription
factors (TFs) that are potential regulators of a set of co-regulated or
co-expressed genes. By scanning input sequences (e.g., promoters) against a
database of known TF binding motifs, **PscanR** assesses which motifs are
significantly over- or under- represented and highlights candidate binding
sites.

## Workflow overview:

***Input***: A set of DNA sequences (e.g., promoter regions) and a library of
known transcription factors binding motif, such as Position Weight Matrices
(PWMs) from the JASPAR database.

***Analysis***: Identification of enriched motif and calculation of statistical
significance (Z-scores and P-Values).

***Output***: A ranked list of enriched motif, along with their associated
P-Values and Z-scores.

# The Background Model

In order to assess the significance of the obtained result, a background model
must be computed in the form of a PSMatrixList object. 

In PscanR the input sequences are considered a sample drawn from a universe 
consisting of all promoter sequences available for the species under
investigation. This *universe* serves as the background model for statistical
analyses, ensuring a comprehensive and unbiased assessment of the analysis
outcomes.

Because the background is the distribution of scores over the whole promoter
universe, the foreground and the background must describe the *same* kind of
sequence. A promoter window of 200 bp upstream and 50 bp downstream of the TSS
does not yield scores comparable with one of 950 bp upstream and 50 bp
downstream: a longer window offers a motif more places to match, so its best
score rises even when nothing is enriched. The foreground promoters must
therefore be extracted with exactly the window the background was built on, and
each shipped background records that window in its name, as in `950u_50d`.

The next section shows how to use one of the pre-computed backgrounds available
for PscanR. Please refer to the vignette *Building a background for PscanR* 
for a walkthrough on how to build and use a custom background.

## Obtaining a Pre-Computed Background

PscanR ships with precomputed backgrounds for five model organisms: *H. sapiens,
M. musculus, D. melanogaster, A. thaliana and S. cerevisiae.*

`get_availableBG()` lists what is available, and
`generate_psmatrixlist_from_background()` retrieves one and returns it as the
`PSMatrixList` that `pscan()` takes:

```{r list-backgrounds, eval=FALSE}
get_availableBG()
get_availableBG(details = TRUE) # the full version-2 catalog
```

```{r download-background, eval=FALSE}
J2020_PSBG <- generate_psmatrixlist_from_background(
    "Jaspar2020", "hs", c(-200, 50), "hg38"
)
```

Four things identify a background, and they are the four arguments above: the
JASPAR release the motifs come from, the species, the promoter window relative
to the TSS, and the genome assembly. The assembly is required only for *Mus
musculus* and *Homo sapiens*, which have more than one. Available combinations
are:

-   *Homo sapiens*, assemblies hg38 and hs1

-   *Mus musculus*, assemblies mm10 and mm39

-   *Arabidopsis thaliana*, TAIR9

-   *Drosophila melanogaster*, dm6

-   *Saccharomyces cerevisiae*, sacCer3

with promoter windows of 200 bp upstream and 50 bp downstream, 450 upstream and
50 downstream, 500 upstream and 0 downstream, 950 upstream and 50 downstream,
and 1000 upstream and 0 downstream, for the JASPAR 2020, 2022 and 2024
releases. Only JASPAR **core** motifs are used.

`generate_psmatrixlist_from_background()` resolves the resource through 
ExperimentHub, and falls back to the immutable Zenodo record
<https://doi.org/10.5281/zenodo.21821764> with a warning if the Hub cannot be
reached. Both routes serve the same archive and every file is verified by SHA-256 
on arrival. Pass `source = "zenodo"` to go to Zenodo directly.

For a self-contained example, the package ships the JASPAR2020 core matrices
and the matching human 200-up, 50-down background as a table, which
`ps_retrieve_bg_from_file()` combines into the same kind of object without any
download:

```{r offline-background}
J2020 <- readRDS(system.file("extdata", "J2020.rds", package = "PscanR"))
bg_path <- system.file(
    "extdata", "J2020_hg38_200u_50d_UCSC.psbg.txt",
    package = "PscanR"
)
J2020_PSBG <- ps_retrieve_bg_from_file(bg_path, J2020)
J2020_PSBG[[4]] # one matrix, with its background statistics
```

When no shipped background matches the species, window or motif collection an 
analysis needs, one can be built instead; the vignette *Building a background for 
PscanR* covers that case.

# The Input Transcript For `pscan()`

**`pscan()`** takes two arguments: a `DNAStringSet` of promoter sequences, and
the background `PSMatrixList` obtained above. The sections below show 
an example on how to build the promoter sequences, starting from the case where 
it is almost immediate and moving to the one that most often arises in practice.

For humans, mice, and Drosophila, the available background models are based on
RefSeq identifiers, for Arabidopsis on TAIR, and for yeast on SGD IDs. Other
types of gene identifiers can also be used if the corresponding background model
is created; the *Building a background for PscanR* vignette covers that case.

## Scanning from a list of transcripts

PscanR works at the promoter level, and a transcript identifier is an
unambiguous match for one promoter within a given assembly and annotation. A
list of transcripts therefore determines a promoter set outright, and nothing
has to be decided.

The package ships an example of such a list: human NRF1 target transcripts 
in RefSeq, curated from the literature.

```{r chunk-2}
file_path <- system.file("extdata", "nrf1100.txt", package = "PscanR")
target <- read.csv(file_path, header = FALSE)
head(target)
```

The package also ships the matching promoter sequences, human, 200 bp upstream 
and 50 bp downstream of the TSS, which is the window of the background loaded 
above:

```{r load-promoters}
library(Biostrings)
prom_seq <- readRDS(
    system.file("extdata", "prom_seq.rds", package = "PscanR")
)
length(prom_seq)
unique(width(prom_seq)) # 200 upstream + 50 downstream
head(names(prom_seq), 3)
```

Both name transcripts the same way, down to the RefSeq version, so the
foreground is one subset:

```{r match-by-name}
foreground <- prom_seq[names(prom_seq) %in% target[[1]]]
length(foreground)
```

Every transcript in the list has a promoter in the bundled set, so all 90 are
kept here; the same expression is what selects a foreground out of a promoter
set covering a whole genome. Scanning that foreground against the background is
the whole of the analysis:

```{r run-pscan}
res <- pscan(foreground, J2020_PSBG, BPPARAM = BiocParallel::SerialParam())
head(ps_results_table(res))
```

`ps_results_table()` orders motifs by increasing p-value. NRF1 comes first, by a
wide margin over the second motif, which is the expected result for a set of
curated NRF1 targets.

## When the list is of genes

A list of transcripts is not what most analyses produce. A differential
expression experiment usually returns **genes**, and a gene list does not determine 
a promoter set: a gene may have several transcription start sites, and therefore
several promoters, whose sequences differ.

One promoter per gene has to be chosen. That choice cannot be avoided, only
made explicitly or left implicit, and leaving it implicit means the result
depends on how finely each gene happens to be annotated.
`ps_select_promoters()` makes it explicitly, by a stated rule, and records for
every gene which transcript was chosen and why.

It is given the genes of interest, the promoters to choose from, and a table
relating one to the other. A few of the genes behind the NRF1 targets stand in 
for what a differential expression analysis would return:

```{r gene-list}
genes <- c(
    "ALAS1", "ARPC1A", "CAPNS1", "CHMP2A",
    "EIF2S1", "GTF2F1", "LONP1", "POLR2A"
)
```

`org.Hs.eg.db` supplies the transcripts of each, which is the relating table.
Note how many there are: eight genes give far more than eight candidate promoters.

```{r build-annotation}
library(org.Hs.eg.db)
annotation <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = genes, keytype = "SYMBOL", columns = "REFSEQ"
)
annotation <- annotation[!is.na(annotation$REFSEQ), ]
nrow(annotation)
table(annotation$SYMBOL)
```

Note also that the identifiers in this table carry no version, whereas the
promoters are named `NM_006559.3` and do. This mismatch is the reason the
function inspects the identifiers before matching them rather than comparing
them as strings. It recognises the RefSeq convention, in which the trailing
number is a release version and two versions name the same transcript, and
normalises the suffix away. It does not do so for TAIR identifiers such as
`AT1G01110.2`, where the same punctuation marks a splice variant that is a
different transcript, frequently with a different start site.

Comparing identifiers as strings is therefore correct for RefSeq and wrong in
general, and deciding which of the two applies is what this function exists to
get right.

```{r select-promoters}
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    gene_col = "SYMBOL", transcript_col = "REFSEQ",
    mode = "representative",
    return = "both"
)
```

The mapping records how each promoter was arrived at, which keeps the choice
auditable rather than implicit. Protein accessions such as `NP_` are discarded
along the way, since they are not transcripts:

```{r selection-mapping}
selection$mapping[, c(
    "gene", "transcript_id", "promoter_id",
    "scheme", "suffix_role", "n_candidates", "decided_by"
)]
```

`ps_selection_summary()` reports the same thing over the whole set, including
how many genes offered a real choice and how many were settled by the tie-break
rather than by the ranking:

```{r selection-summary}
ps_selection_summary(selection$mapping)
```

```{r selection-counts, include=FALSE}
unmapped <- attr(selection$mapping, "ps_unmapped_genes")
n_selected <- length(selection$sequences)
```

Of the `r length(genes)` genes, `r length(unmapped)` is unmapped
(`r paste(unmapped, collapse = ", ")`), because none of its transcripts has a
promoter in this small set. The remaining `r n_selected` are returned as a
`DNAStringSet`, named by gene and transcript so that a result can be traced
back to both:

```{r selected-sequences}
selection$sequences
```

`mode = "representative"` ranks transcripts by the scheme's own rule, which for
RefSeq prefers `NM_` over `NR_` over the predicted `XM_` and `XR_` accessions.
For human hg38 a better ranking is available, because MANE Select and RefSeq
Select name one transcript per gene by curation. `mode = "select"` uses it:

```{r mane-select, eval=FALSE}
select_tx <- ps_load_select_transcripts("hg38")
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    select_transcripts = select_tx,
    mode = "select",
    return = "both"
)
```

`ps_load_select_transcripts()` retrieves the two tracks from the UCSC API, which
takes around a minute and requires a third-party service to be reachable;
so it is not evaluated in the vignette build. Run once, the table is cached under
`tools::R_user_dir("PscanR", "cache")` and later calls read it from there. The
metadata is specific to human hg38, so for other organisms either
`mode = "representative"` applies, or a species-specific table is supplied
through `select_transcripts`.

Either way the result is a `DNAStringSet` of one promoter per gene, which is
passed to `pscan()` exactly as the foreground of the previous section was. Both
it and the background must describe the same window, here 200 bp upstream and
50 bp downstream. Had the background been a `950u_50d` one, these promoters
would have had to be extracted again over 950 bp upstream rather than 200,
because a longer window gives a motif more places to match and raises its best
score whether or not anything is enriched.

# Choosing a `BPPARAM`

`pscan()` parallelises over motifs through **BiocParallel**, and the `BPPARAM`
argument selects the strategy. The recommendation is:

-   **`MulticoreParam(n)` on Linux and macOS.** It parallelises by forking, which
    is the faster of the two because the workers inherit the session's memory
    instead of receiving a copy of it.

-   **`SnowParam(n)` where forking is unavailable**, which in practice means
    Windows. It starts independent processes and communicates over sockets, so
    it works everywhere at the cost of the data transfer.

-   **`SerialParam()` when debugging**, since it starts no workers at all and a
    failure therefore surfaces in the session rather than inside one of them.

The default is `bpparam()`, which picks a strategy and a worker count from the
session's own configuration.

Raise the number of cores when a scan on one appears slow. More cores may not
help on a scan that is already quick, because starting the workers and shipping
the matrices out to them costs time that a short scan does not repay.

# Session information

```{r session-info}
sessionInfo()
```
