---
title: "PscanR Concepts and Input Preparation"
author:
    - Diana Betelli
    - Federico Zambelli
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
vignette: >
    %\VignetteIndexEntry{2. PscanR Concepts and Input Preparation}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    markdown:
        wrap: 80
---

```{r setup, include = FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Introduction

PscanR is an R implementation of the algorithm developed by
[Zambelli et al. (2009)](https://doi.org/10.1093/nar/gkp464) for analysing
sequence motif enrichment. It identifies transcription
factors (TFs) that are potential regulators of a set of co-regulated or
co-expressed genes. By scanning input sequences (e.g., promoters) against a
database of known TF binding motifs, PscanR tests for motif enrichment and
reports candidate binding sites. Negative z-scores indicate lower mean scores
than the background; the reported upper-tail p-values test enrichment, not
depletion. Motif matches alone do not establish transcription factor binding.

## Workflow overview

**Input:** DNA sequences supplied as a `DNAStringSet`, and transcription factor
binding matrices with matching background statistics supplied as a
`PSMatrixList`.

**Analysis:** Calculation of the best motif match in each promoter, followed by
comparison of foreground and background scores.

**Output:** Motif-level z-scores, p-values, and adjusted p-values, together with
the best match in each promoter for each motif.

# The background model

Statistical assessment requires a background model, represented by background
statistics stored in a `PSMatrixList` object.

PscanR treats the input sequences as a sample from a defined reference
population, usually the annotated promoters of the species under investigation.
This reference population supplies the background model for statistical
analyses. The choice of this reference population affects interpretation;
using all annotated promoters does not eliminate sequence-composition or
gene-selection biases.

Because the background is the distribution of scores over the whole promoter
universe, the foreground and the background must describe the *same* kind of
sequence. A promoter window of 200 bp upstream and 50 bp downstream of the
transcription start site (TSS)
does not yield scores comparable with one of 950 bp upstream and 50 bp
downstream: a longer window offers more candidate matches and can therefore
increase the best score without biological enrichment. Extract foreground
promoters using the background's window and compatible reference sequences and
annotations. Each precomputed background records its window in its name, as in
`950u_50d`.

The next section demonstrates the use of a precomputed background. The vignette
[Building a background for PscanR](PscanR_Custom_Background.html) describes the construction and use of custom
backgrounds.

## Obtaining a precomputed background

PscanR provides access to precomputed backgrounds for five model organisms:
*Homo sapiens*, *Mus musculus*, *Drosophila melanogaster*,
*Arabidopsis thaliana*, and *Saccharomyces cerevisiae*.

`get_availableBG()` lists the available backgrounds, and
`generate_psmatrixlist_from_background()` retrieves the selected background as
a `PSMatrixList` for use with `pscan()`:

```{r list-backgrounds, eval=FALSE}
get_availableBG()
get_availableBG(details = TRUE) # the full version-2 catalog
```

```{r download-background, eval=FALSE}
J2020_PSBG <- generate_psmatrixlist_from_background(
    "Jaspar2020", "hs", c(-200, 50), "hg38"
)
```

The four arguments above identify the JASPAR release, species, promoter window
relative to the TSS, and genome assembly. The
assembly must be specified for mouse and human, for which multiple assemblies
are available. The supported species and assemblies are:

-   *Homo sapiens*, assemblies hg38 and hs1

-   *Mus musculus*, assemblies mm10 and mm39

-   *Arabidopsis thaliana*, TAIR9

-   *Drosophila melanogaster*, dm6

-   *Saccharomyces cerevisiae*, sacCer3

The available promoter windows span 200 bp upstream and 50 bp downstream, 450 upstream and
50 downstream, 500 upstream and 0 downstream, 950 upstream and 50 downstream,
and 1000 upstream and 0 downstream, for the JASPAR 2020, 2022, and 2024
releases. Only JASPAR CORE motifs are used.

`generate_psmatrixlist_from_background()` resolves the resource through
ExperimentHub, and falls back to the immutable Zenodo record
<https://doi.org/10.5281/zenodo.21821764> with a warning if the Hub cannot be
reached. Both routes serve the same archived resources, and downloaded files
are verified against their SHA-256 checksums. Set `source = "zenodo"` to retrieve
the resource directly from Zenodo.

For an offline example, the package includes JASPAR 2020 CORE matrices and a
matching human background table for 200 bp upstream and 50 bp downstream of the
TSS. `ps_retrieve_bg_from_file()` combines them into a `PSMatrixList`:

```{r offline-background}
J2020 <- readRDS(system.file("extdata", "J2020.rds", package = "PscanR"))
bg_path <- system.file(
    "extdata", "J2020_hg38_200u_50d_UCSC.psbg.txt",
    package = "PscanR"
)
J2020_PSBG <- ps_retrieve_bg_from_file(bg_path, J2020)
J2020_PSBG[[4]] # one matrix, with its background statistics
```

If no available background matches the required species, window, or motif
collection, construct a custom background as described in [Building a background for PscanR](PscanR_Custom_Background.html).

# Preparing promoter sequences for `pscan()`

The two principal inputs to `pscan()` are a `DNAStringSet` of promoter sequences
and a background `PSMatrixList`. The following sections demonstrate promoter
selection from transcript identifiers and from gene identifiers.

For humans, mice, and Drosophila, the available background models are based on
RefSeq identifiers; for *Arabidopsis*, on TAIR; and for yeast, on SGD. Other
types of gene identifiers can also be mapped to the annotation used by an
existing background. Custom backgrounds can use other annotations, as described
in [Building a background for PscanR](PscanR_Custom_Background.html).

## Scanning from a list of transcripts

PscanR operates on promoter sequences. When transcript identifiers map uniquely
within the chosen assembly and annotation, their corresponding promoters can
be selected directly without choosing among transcripts of the same gene.

The package includes an example list of human NRF1 target transcripts
in RefSeq, curated from the literature.

```{r chunk-2}
file_path <- system.file("extdata", "nrf1100.txt", package = "PscanR")
target <- read.csv(file_path, header = FALSE)
head(target)
```

The bundled human promoter sequences span 200 bp upstream and 50 bp downstream
of the TSS, matching the background loaded above:

```{r load-promoters}
library(Biostrings)
prom_seq <- readRDS(
    system.file("extdata", "prom_seq.rds", package = "PscanR")
)
length(prom_seq)
unique(width(prom_seq)) # 200 upstream + 50 downstream
head(names(prom_seq), 3)
```

The transcript list and promoter names use identical RefSeq identifiers,
including version suffixes. The foreground can therefore be selected by exact
name matching:

```{r match-by-name}
foreground <- prom_seq[names(prom_seq) %in% target[[1]]]
length(foreground)
```

Every transcript in the list has a promoter in the bundled set, so all 90 are
retained. The same operation can select a foreground from a genome-wide
promoter set. The selected sequences are then scanned against the background:

```{r run-pscan}
res <- pscan(foreground, J2020_PSBG, BPPARAM = BiocParallel::SerialParam())
head(ps_results_table(res))
```

`ps_results_table()` returns motifs ordered by increasing p-value, with
decreasing z-score used to resolve ties. NRF1 is the highest-ranked motif,
consistent with the origin of the curated target set.

## Selecting promoters from a gene list

A differential expression analysis often produces a gene list. A gene may have
several transcription start sites and therefore several distinct promoters, so
gene identifiers alone do not uniquely specify a promoter set.

This workflow selects one promoter per gene to avoid giving greater weight to
genes with more annotated transcripts. `ps_select_promoters()` applies an
explicit selection rule and records the selected transcript and the reason for
its selection. This provides reproducibility without establishing which
transcript is active in the experimental condition.

The function requires the genes of interest, candidate promoter sequences, and
a gene-to-transcript annotation table. The following subset of genes from the
NRF1 target list illustrates the input format:

```{r gene-list}
genes <- c(
    "ALAS1", "ARPC1A", "CAPNS1", "CHMP2A",
    "EIF2S1", "GTF2F1", "LONP1", "POLR2A"
)
```

`org.Hs.eg.db` supplies the gene-to-transcript mapping. These eight genes have
more than eight candidate transcripts:

```{r build-annotation}
library(org.Hs.eg.db)
annotation <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = genes, keytype = "SYMBOL", columns = "REFSEQ"
)
annotation <- annotation[!is.na(annotation$REFSEQ), ]
nrow(annotation)
table(annotation$SYMBOL)
```

The annotation table contains unversioned RefSeq accessions, whereas promoter
names such as `NM_006559.3` include a version suffix. The function recognises
the RefSeq convention and removes the suffix for identifier matching. It
preserves the suffix in TAIR identifiers such as `AT1G01110.2`, where it
distinguishes transcript variants that may have different start sites.

Thus, RefSeq matching uses the accession after version normalisation, whereas
TAIR matching uses the complete transcript identifier. Matching RefSeq versions
does not imply that their sequences or annotated start sites are identical;
the promoter sequences must still correspond to the chosen reference.

```{r select-promoters}
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    gene_col = "SYMBOL", transcript_col = "REFSEQ",
    mode = "representative",
    return = "both"
)
```

The returned mapping documents each selection. Protein accessions such as
`NP_` are excluded because they do not identify transcripts:

```{r selection-mapping}
selection$mapping[, c(
    "gene", "transcript_id", "promoter_id",
    "scheme", "suffix_role", "n_candidates", "decided_by"
)]
```

`ps_selection_summary()` summarises the mapping, including the number of genes
with multiple candidates and the number requiring a tie-breaking rule after
transcript ranking:

```{r selection-summary}
ps_selection_summary(selection$mapping)
```

```{r selection-counts, include=FALSE}
unmapped <- attr(selection$mapping, "ps_unmapped_genes")
n_selected <- length(selection$sequences)
```

The number of unmapped genes is `r length(unmapped)` out of `r length(genes)`;
the unmapped identifiers are `r paste(unmapped, collapse = ", ")`. These genes
have no matching transcript promoter in the bundled set. The
`r n_selected` selected promoters are returned as a
`DNAStringSet`, named by gene and transcript so that a result can be traced
back to both:

```{r selected-sequences}
selection$sequences
```

`mode = "representative"` ranks transcripts by the identifier scheme's rule.
For RefSeq, it prefers `NM_` over `NR_`, followed by predicted `XM_` and `XR_`
accessions.
For human hg38, MANE Select and RefSeq Select provide curated transcript
preferences. These can be used with `mode = "select"`:

```{r mane-select, eval=FALSE}
select_tx <- ps_load_select_transcripts("hg38")
selection <- ps_select_promoters(
    genes = genes,
    promoter_sequences = prom_seq,
    annotation = annotation,
    select_transcripts = select_tx,
    mode = "select",
    return = "both"
)
```

`ps_load_select_transcripts()` retrieves the two tracks through the UCSC API.
This network-dependent example is not evaluated during vignette compilation.
After retrieval, the table is cached under
`tools::R_user_dir("PscanR", "cache")` for subsequent calls. The
metadata is specific to human hg38, so for other organisms either
`mode = "representative"` applies, or a species-specific table is supplied
through `select_transcripts`.

With `return = "both"`, the selected `DNAStringSet` is available as
`selection$sequences` and can be passed to `pscan()`. It contains one promoter
per mapped gene. The sequences and background must use the same window, here
200 bp upstream and 50 bp downstream. A `950u_50d` background would require
promoters extracted over the corresponding interval.

# Configuring parallel execution

`pscan()` parallelises over motifs through **BiocParallel**, and the `BPPARAM`
argument selects the execution strategy:

-   **`MulticoreParam(n)` on systems supporting forked execution.** Workers
    inherit the parent process's memory, which can reduce data-transfer
    overhead on Linux and macOS. Suitability also depends on the execution
    environment; forked execution is unavailable on Windows.

-   **`SnowParam(n)` for portable parallel execution.** Independent R processes
    communicate through sockets on Windows and Unix-like systems. Worker
    startup and data transfer introduce additional overhead.

-   **`SerialParam()` for debugging and small examples.** Computation occurs
    within the current R process, simplifying error inspection.

The default is `bpparam()`, which selects a strategy and worker count from the
session configuration.

Select the worker count according to the workload and available resources.
For short scans, worker startup and data-transfer costs may exceed the benefit
of parallel execution.

# Session information

```{r session-info}
sessionInfo()
```
