---
title: "PscanR Concepts and Input Preparation"
author:
    - Diana Betelli
    - Federico Zambelli
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
        toc_depth: 2
vignette: >
    %\VignetteIndexEntry{2. PscanR Concepts and Input Preparation}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    markdown:
        wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

**PscanR** is an R implementation of the algorithm developed by Zambelli et al.
(2009) for analyzing sequence motif enrichment. It helps identify transcription
factors (TFs) that are potential regulators of a set of co-regulated or
co-expressed genes. By scanning input sequences (e.g., promoters) against a
database of known TF binding motifs, **PscanR** assesses which motifs are
significantly over- or under- represented and highlights candidate binding
sites.

## Workflow overview:

***Input***: A set of DNA sequences (e.g., promoter regions) and a library of
known transcription factors binding motif, such as Position Weight Matrices
(PWMs) from the JASPAR database.

***Analysis***: Identification of enriched motif and calculation of statistical
significance (Z-scores and P-Values).

***Output***: A ranked list of enriched motif, along with their associated
P-Values and Z-scores.

# The Background Model

In order to assess the significance of the obtained result, a background model
must be computed in the form of a PSMatrixList object. In other methods the
background is generated by shuffling the columns of the motif or by building
random sequence sets of the same size and length of the sequence set
investigated.

In PscanR, instead, the input sequences are considered a sample drawn from a
universe consisting of all promoter sequences available for the species under
investigation. This *universe* serves as the background model for statistical
analyses, ensuring a comprehensive and unbiased assessment of the analysis
outcomes.

Because the background is the distribution of scores over the whole promoter
universe, the foreground and the background must describe the *same* kind of
sequence. A promoter window of 200 bp upstream and 50 bp downstream of the TSS
does not yield scores comparable with one of 950 bp upstream and 50 bp
downstream: a longer window offers a motif more places to match, so its best
score rises even when nothing is enriched. The foreground promoters must
therefore be extracted with exactly the window the background was built on, and
each shipped background records that window in its name, as in `950u_50d`.

The remainder of this vignette covers the other half of preparing an input,
which is arriving at the transcript identifiers that name the promoters to be
scanned.

# The Input Transcript For `pscan()`

**`pscan()`** accepts as input a DNAStringSet object containing the DNA
sequences of co-regulated or co-expressed genes. This object can be built from a
list of gene/transcript identifiers.

For humans, mice, and Drosophila, the available background models are based on
RefSeq identifiers, for Arabidopsis on TAIR, and for yeast on SGD IDs. Other
types of gene identifiers can also be used if the corresponding background model
is created; the *Building a background for PscanR* vignette covers that case.

Below is an example dataset containing **human NRF1 target transcripts**
(RefSeq), curated from the literature. Since PscanR works at the transcript
level, gene identifiers must be converted to their corresponding transcript IDs
before analysis. The number **100** indicates that 100% transcripts in the list
correspond to known NRF1 target.

```{r chunk-2}
file_path <- system.file("extdata", "nrf1100.txt", 
                            package = "PscanR")
target <- read.csv(file_path, header = FALSE)
head(target)
```

## Gene IDs - Transcript IDs conversion

Since the functions work with Transcript IDs, we provide different pipelines for
various organism to help users in the conversion from Gene IDs.

### ENSEMBL

For ENSEMBL-based annotation, the `biomaRt` package can be used. Below is an
example for *Homo sapiens*.

```{r eval=FALSE, include=TRUE}
# Homo sapiens example 
mart <- biomaRt::useMart('ensembl', dataset = 'hsapiens_gene_ensembl')
GeneTranscript_ids <- biomaRt::getBM(attributes = 
                                        c('ensembl_gene_id',
                                            'ensembl_transcript_id',
                                            'transcript_tsl'),
                                        mart = mart) 
```

```{r eval = TRUE, include = FALSE}
path <- system.file('extdata', 'GeneTranscript_ids.rds', package = 'PscanR')
GeneTranscript_ids <- readRDS(path)
```

```{r eval = TRUE, include = TRUE}
head(GeneTranscript_ids)
```

This code chunk retrieves all gene identifiers for *Homo sapiens*, along with
the corresponding transcript IDs and their Transcript Support Level (TSL).

To slect only trasncripts with the highest level of experimental support, filter
by TSL1. This means that there is experimentally evidence that all the splice
junctions of the annotated transcript actually exist in the cells:

```{r chunk-3}
genes <- c('ENSG00000271254', 'ENSG00000277196', 'ENSG00000277196')

transcripts <- GeneTranscript_ids[
    GeneTranscript_ids[, 1] %in% genes &
    GeneTranscript_ids[, 3] == 'tsl1',
    2
]

transcripts
```

Change the dataset parameter according to the organism of study:

-   *Mus musculus*: 'mmusculus_gene_ensembl'

-   *Drosophila melanogaster*: 'dmelanogaster_gene_ensembl'. Note that for this
    organism the TSL attribute is not available.

### RefSeq

For RefSeq-based annotations, the AnnotationDbi package can be used together
with organism specific packages. The following is an example with `org.Dm.eg.db` for
*Drosophila melanogaster*.

```{r chunk-4}
library(org.Dm.eg.db)
library(dplyr)

genes <- c('CG46191', 'dbr', 'Pp1-Y2')

transcripts <- AnnotationDbi::select(org.Dm.eg.db,
                                        keys = genes, 
                                        keytype = 'SYMBOL', # Entrez can be used
    columns = 'REFSEQ') %>%
    filter(grepl('^NM_', REFSEQ))

transcripts
```

For other organisms:

-   *Homo sapiens*: use the `org.Hs.eg.db` library

-   *Mus musculus*: use the `org.Mm.eg.db` library

### GENCODE

For GENCODE-based annotations the GFF files must be downloaded directly,
because no R package performs the conversion. The `biomaRt` package is an
alternative, but it returns ENSEMBL identifiers, which carry no variant
information.

Below is an example with *Homo sapiens*. Because of high computational time and
memory requirements, it is not run during the building of the vignette.

```{r eval=FALSE, include=TRUE}
gtf <- rtracklayer::import("gencode.v47.basic.annotation.gff3.gz")

genes <- c(
    'ENSG00000171723.16', 'ENSG00000112715.26',
    'ENSG00000142156.16', 'ENSG00000170873.19'
)

transcripts <- gtf[
    gtf$gene_id %in% genes &
    !is.na(gtf$transcript_support_level) &
    gtf$transcript_support_level == '1',
]$transcript_id

transcripts <- unique(transcripts)
```

```{r eval=TRUE, include=FALSE}
transcripts <- readRDS(
    system.file('extdata', 'transcripts.rds', package = 'PscanR')
)
```

```{r chunk-5}
transcripts 
```

### SGD

In *Saccharomyces cerevisiae*, genes are often referred to by their **systematic
ORF IDs** (e.g., `YPR195C`). **Gene symbols** can be converted to them as shown
below. Note that not all the systematic IDs have a corresponding symbol.

```{r chunk-6}
library(org.Sc.sgd.db)
gene_symbols <- c('TIF5', 'CDC28', 'RPL3')

ORF <- AnnotationDbi::select(org.Sc.sgd.db,
                                keys = gene_symbols,
                                columns = c('ORF', 'GENENAME'),
                                keytype = 'GENENAME')

ORF # take the second column for systematic ids
```

# Choosing a `BPPARAM`

`pscan()` parallelises over motifs through **BiocParallel**, and the `BPPARAM`
argument selects the strategy. The recommendation is:

-   **`MulticoreParam(n)` on Linux and macOS.** It parallelises by forking, which
    is the faster of the two because the workers inherit the session's memory
    instead of receiving a copy of it.

-   **`SnowParam(n)` where forking is unavailable**, which in practice means
    Windows. It starts independent processes and communicates over sockets, so
    it works everywhere at the cost of the data transfer.

-   **`SerialParam()` when a run must be exactly reproducible** or when
    debugging, since it starts no workers at all.

The default is `bpparam()`, which picks a strategy and a worker count from the
session's own configuration.

More cores are not always better. Both parallel backends pay a fixed cost to
start workers and to ship the matrices out to them, and that cost is repaid only
if each worker then has enough scanning to do. A foreground of a few hundred
promoters can therefore be *slower* on eight cores than on one, while a
background build over an entire promoter universe scales well. Both worked
examples in this collection quote measured timings for their own scans, and the
custom-background vignette compares a one-core and an eight-core build of the
same background directly.

# Session information

```{r session-info}
sessionInfo()
```
