---
title: "Motif enrichment analysis with PscanR: mouse retinal promoters"
author:
  - Federico Zambelli
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{3. Motif enrichment analysis with PscanR: mouse retinal promoters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
library(PscanR)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
)

fixture <- function(...) {
    system.file(
        "extdata", "vignettes", "mouse_hd_retina", ...,
        package = "PscanR", mustWork = TRUE
    )
}
```

# Overview

PscanR looks for transcription factor binding motifs that are over-represented
in the regulatory regions of a set of genes. The idea in brief: take the
promoters of genes that behave alike (co-expressed, co-regulated, differentially
expressed in the same contrast), scan them with a collection of position
frequency matrices, and ask which motifs score higher there than they do across
all promoters of the organism, that is the universe background set.

The package ships a collection of precomputed background sets for five model
organisms: human, mouse and fruit fly (RefSeq annotation), *Arabidopsis* (TAIR
annotation), and yeast (SGD annotation). The package also provides functions to
compute and use custom backgrounds, for those or for different organisms (not
covered in this vignette).

This vignette works through a PscanR analysis on mouse promoters and through
several of the functions the package provides for inspecting the result. 

## The data

The example uses gene expression data from a study of the R6/1 mouse model of
Huntington's disease, which profiled retina and striatum from the same animals.

> Cano-Cano F, Martín-Loro F, Gallardo-Orihuela A, González-Montelongo MC,
> Ortuño-Miquel S, Hervás-Corpión I, de la Villa P, Ramón-Marco L,
> Navarro-Calvo J, Gómez-Jaramillo L, Arroba AI, Valor LM. *Retinal dysfunction
> in Huntington's disease mouse models concurs with local gliosis and microglia
> activation.* Scientific Reports 14:4176 (2024).
> doi:10.1038/s41598-024-54347-8

Differential expression was computed separately in the two tissues, and the
resulting genes split by tissue and direction into six sets. The input is a
two-column table: a gene identifier and the set it belongs to.

Every fixture this vignette reads was produced by `run_uc2_pscan.R`, which
performs the scan, followed by `build_uc2_vignette_data_v3.R`, which condenses
its output into the single object loaded below. Both are kept in
`UseCase/UC2_Mouse_HD_Retina/` of the PscanR source repository rather than
shipped with the package, because they require network access and the full mouse
annotation.

```{r read-gene-sets}
membership <- read.delim(
    fixture("paper_subset_membership.tsv"),
    stringsAsFactors = FALSE
)
str(membership)
gene_sets <- split(membership$ensembl_gene_id, membership$subset)
lengths(gene_sets)
```

```{r subset-key, echo=FALSE}
subset_labels <- c(
    a = "retina-only down", b = "shared down", c = "striatum-only down",
    d = "retina-only up", e = "shared up", f = "striatum-only up"
)
knitr::kable(
    data.frame(
        set = names(subset_labels),
        definition = unname(subset_labels),
        genes = lengths(gene_sets)[names(subset_labels)],
        row.names = NULL
    ),
    caption = "The six gene sets, by tissue and direction of change."
)
```

The rest of this vignette follows set `d`, the
`r length(gene_sets$d)` genes upregulated in retina and not in striatum, and
uses set `f`, the `r length(gene_sets$f)` genes upregulated in striatum and not
in retina, as the comparison.

# The background

PscanR precomputed backgrounds are indexed by organism, genome assembly,
promoter window and motif collection. `get_availableBG()` lists what is
available in the catalog bundled with the package:

```{r discover-background}
catalog <- get_availableBG("J2020_mm10_950u_50d", details = TRUE)
knitr::kable(
    catalog[c(
        "organism", "assembly", "upstream", "downstream",
        "jaspar_release", "motif_count", "promoter_count"
    )],
    row.names = FALSE,
    caption = "The matching precomputed background."
)
```

`generate_psmatrixlist_from_background()` retrieves a background and attaches it
to the matching JASPAR collection, returning a `PSMatrixList`, that is the
object PscanR works with throughout.

```{r retrieve-background, eval=FALSE}
background <- generate_psmatrixlist_from_background(
    JASPAR_matrix = "Jaspar2020",
    org = "mm",
    prom_reg = c(-950, 50),
    assembly = "mm10",
    version = 2
)
length(background)
#> [1] 746
```

That returns the JASPAR 2020 CORE *vertebrates* non-redundant collection, 746
matrices, each carrying three metrics: how many promoters its background was
computed over, and the mean and standard deviation of its score across them.

Take notice, the promoter window selected, 950 bp upstream to 50 bp downstream
of the transcription start site here, is important: **to be comparable, the
foreground promoters must be extracted with the same promoter window the
background was built from**. 

# From genes to promoters

The `pscan()` function of PscanR takes as input a `DNAStringSet` object
containing a set of promoters from co-regulated or co-expressed genes. The user
needs to take care to match the promoter window of the selected background.

> The ideal workflow would require knowing which transcripts are co-regulated or
> co-expressed, in order to retrieve the exact set of promoters regulating their
> expression. Usually, though, the output of a differential expression analysis
> is a list of genes rather than transcripts.
>
> Since a gene may have several transcription start sites and therefore
> different promoters, when working at the gene level the choice of which
> promoter to use for each gene is necessarily arbitrary. This is not ideal, but
> it removes the barrier of having to work at the transcript level, which
> remains the preferred option.

The `ps_select_promoters()` function is a utility function that can be used to
pass from gene or transcript identifiers to their promoter sequences for the
supported organisms and annotations, deterministically resolving a gene list to
one promoter per gene.

Before it can do that it has to work out what the identifiers mean, because the
same punctuation carries different meanings in different references. In RefSeq,
`NM_000546.6` ends in a release **version**, so that two versions name one
transcript. In TAIR, `AT1G01110.2` ends in a **splice variant**, which is a
different transcript, often with a different start site. PscanR detects the convention and reports it.

The gene identifiers in this experiment are Ensembl, so they have to be 
translated to RefSeq through `org.Mm.eg.db`, and the RefSeq version 
`org.Mm.eg.db` may not be the version the UCSC promoter track carries. 
Under the RefSeq rule the two still match, because the version is normalised away.

The following example uses ten real multi-transcript genes from this dataset,
shipped with the example as a small fixture:

```{r promoter-scheme, message=TRUE}
example <- readRDS(fixture("promoter_selection_example.rds"))

selection <- ps_select_promoters(
    genes = example$genes,
    promoter_sequences = example$promoter_sequences,
    annotation = example$annotation,
    gene_col = "ENSEMBL", transcript_col = "REFSEQ",
    mode = "representative",
    return = "mapping",
    sequence_names = "gene_transcript"
)
```

The returned mapping records the reason for the choice:

```{r promoter-scheme-table, echo=FALSE}
knitr::kable(
    selection[, c(
        "gene", "promoter_id", "scheme", "suffix_role",
        "transcript_class", "selection_source", "n_candidates", "decided_by"
    )],
    row.names = FALSE,
    caption = "One promoter per gene, and why each was chosen."
)
```

For RefSeq the rule is to prefer an `NM_` messenger transcript over an `NR_`
non-coding one, and to break any remaining tie by accession order. Protein
accessions (`NP_`, `XP_`) are excluded before ranking.

For our dataset we need promoter sequences matching the background selected
above: mm10 assembly, the UCSC `ncbiRefSeqCurated` track, 950 bp upstream to 50
bp downstream of the TSS.

The following chunk shows how to do it but is not evaluated during the vignette
build, since it requires external annotation and genome packages. Please notice
it needs the `AnnotationDbi`, `Biostrings`, `GenomicRanges`, `IRanges`,
`org.Mm.eg.db` and `BSgenome.Mmusculus.UCSC.mm10` packages.

`mm10_950u_50d_refseq_snapshot.rds` is shipped with this example and records the
transcript identifiers and TSS coordinates of the promoter universe the
background was computed over.


```{r select-real, eval=FALSE}

retina_up_genes <- gene_sets$d

# The transcripts and start sites the background was built from.
snapshot <- readRDS(fixture("mm10_950u_50d_refseq_snapshot.rds"))
transcripts <- snapshot$transcripts

transcript_tss <- GenomicRanges::GRanges(
    seqnames = transcripts$seqname,
    ranges = IRanges::IRanges(start = transcripts$tss, width = 1L),
    strand = transcripts$strand
)
names(transcript_tss) <- transcripts$transcript_id

# promoters() takes the region around each 5' end, so it is strand-aware and
# anchored on the transcription start site.
ranges <- GenomicRanges::promoters(
    transcript_tss,
    upstream = snapshot$upstream,
    downstream = snapshot$downstream
)

genome <- BSgenome.Mmusculus.UCSC.mm10::BSgenome.Mmusculus.UCSC.mm10
promoters_mm10 <- Biostrings::getSeq(genome, ranges)
names(promoters_mm10) <- names(ranges)

# Ensembl gene identifiers to RefSeq transcripts. This returns several rows per
# gene, including the protein accessions ps_select_promoters() discards.
annotation <- AnnotationDbi::select(
    org.Mm.eg.db::org.Mm.eg.db,
    keys = retina_up_genes,
    keytype = "ENSEMBL",
    columns = c("REFSEQ", "SYMBOL")
)

retina_mapping <- ps_select_promoters(
    genes = retina_up_genes,
    promoter_sequences = promoters_mm10,
    annotation = annotation,
    gene_col = "ENSEMBL", transcript_col = "REFSEQ",
    mode = "representative",
    return = "mapping",
    sequence_names = "gene_transcript"
)

retina_up_promoters <- promoters_mm10[retina_mapping$promoter_id]
```

That mapping was saved along with the scan it fed, so the rest of this vignette
can read it directly from the precomputed object.

```{r load-precomputed}
precomputed <- readRDS(fixture("uc2_vignette_data.rds"))
retina_mapping <- precomputed$retina_mapping
```

`ps_selection_summary()` reports how many genes were asked for, how many were placed, and how often
there was a real choice to make.

```{r selection-summary}
ps_selection_summary(retina_mapping)$overall
```

Nearly half the genes had more than one candidate promoter, and most of those
were settled by the tie-break rule: for a gene with several curated 
messenger transcripts there is no principled way to pick one, 
only a deterministic one.

The genes that go unmapped are those with no curated RefSeq transcript to map
to: most have no RefSeq accession at all in `org.Mm.eg.db`, three are
mitochondrial genes that carry only a protein accession and are outside the
nuclear promoter set anyway, and one has only predicted `XM_` models, which the
curated track does not include. They are listed on the mapping rather than
dropped silently:

```{r unmapped}
head(attr(retina_mapping, "ps_unmapped_genes"), 4)
```

`selection_source` names the rule that decided each promoter:

```{r selection-by-source}
ps_selection_summary(retina_mapping)$by_source
```

# Running the scan

`pscan()` takes the foreground promoter sequences and the background
`PSMatrixList`, and returns the updated `PSMatrixList` carrying the
per-sequence results for every motif.

The BPPARAM from `BiocParallel` can be used to set the level of parallelism:

- `SerialParam()` uses one R process and is the safest choice for package checks
  and reproducible debugging.

- `SnowParam()` starts independent socket-based R workers; it is the portable
  parallel choice on Windows and Unix-like systems, but has greater worker
  startup and memory overhead.

- `MulticoreParam()` uses lower-overhead forked workers and is generally
  preferable for local parallel execution on Linux or other supported Unix-like
  systems; it should not be used on Windows.

The next chunk is not evaluated by default to limit the vignette's build time,
but the user can run it. Scanning the two sets used here took
`r round(sum(precomputed$timings$elapsed_seconds[precomputed$timings$subset %in% c("d", "f")]), 0)`
seconds with `BPPARAM` set to
`BiocParallel::MulticoreParam(workers = 8)`.

```{r scan, eval=FALSE}

BPPARAM <- BiocParallel::SerialParam() # Serial - Compatible but slower

# Cross-platform parallel execution with four independent R workers:
# BPPARAM <- BiocParallel::SnowParam(workers = 4, type = "SOCK") #Use this on Windows

# Lower-overhead forked execution on supported Unix-like systems: #Use this on Linux and MacOS
# BPPARAM <- BiocParallel::MulticoreParam(workers = 4)

results <- pscan(
    retina_up_promoters, background,
    BPPARAM = BPPARAM
)
```

`ps_results_table()` turns that object into the enrichment table.

```{r results-table, eval=FALSE}
enrichment <- ps_results_table(results)
```

To shorten the vignette's build time the scan was saved into the same RDS file
the mapping came from. The objects below are that saved result; everything from
here on is computed from them.

```{r load-results}
# Exactly what ps_results_table() returned for the retina-only upregulated
# promoters, and for the striatum-only comparison.
enrichment <- precomputed$enrichment_d
dim(enrichment)

# The same scan kept as a PSMatrixList rather than a table, so the
# per-promoter detail survives, restricted to a few motifs to keep it small.
results_subset <- precomputed$results_subset_d
results_subset
```

Those motifs were chosen to span what the scan reports: the interferon
regulatory factors and STATs the original study is about, the GC-rich motifs
that turn out to top the table, and one depleted motif for contrast.

```{r subset-names}
TFBSTools::name(results_subset)
```

The enrichment table is one row per motif, keyed by JASPAR matrix identifier:

```{r results-table-shown}
head(enrichment[order(-enrichment$ZSCORE), ], 8)
```

The top of the full table is occupied not by the interferon factors but by
GC-rich motifs: KLF15, ZBTB14, the SP and EGR zinc fingers. IRF3, the strongest
member of the family the study reported, ranks
`r which(rownames(enrichment[order(-enrichment$ZSCORE), ]) == "MA1418.1")`th
with a z-score of
`r sprintf("%.2f", enrichment["MA1418.1", "ZSCORE"])`. 

The background parameters travel with the result, so they can be read back from
it at any point with `ps_get_bg_table()`:

```{r bg-table}
ps_get_bg_table(results_subset)
```

`BG_SIZE` is the number of mouse promoters the background was computed over;
`BG_MEAN` and `BG_STDEV` describe how that motif scores across them.

Each row of the enrichment table compares the foreground mean score against the
background mean, turns the difference into a z-score, and reports the
corresponding p-value with a Benjamini–Hochberg FDR across the collection.

`ps_motif_barplot()` ranks the motifs and draws them; the `n` argument limits
the plot to the top of the list (default to at most 20). `group = "class"`
colours each bar by the structural class of the factor, which
`ps_motif_class()` reads off the JASPAR matrices.

```{r top-motifs, fig.width=8, fig.height=5, dpi=120}

table(ps_motif_class(results_subset))

ps_motif_barplot(results_subset, group = "class")
```

Within the subset the C2H2 zinc fingers rank above the interferon factors, as
they do in the full table above. NEUROD2 is there as a reminder that the
z-score is signed: a motif can score well *below* background across a set of
promoters, which can be biologically informative in its own right. The
*P*-value will not report it, because it is a one-sided upper tail. NEUROD2's
z-score of
`r sprintf("%.1f", enrichment["MA0668.1", "ZSCORE"])`
comes with *P* ≈ 1, thus the depletion is visible in the z-score alone.

The function returns a `ggplot` object rather than drawing immediately, so a
title, a different theme or any other layer can be added afterwards:

```{r barplot-composable, eval=FALSE}
ps_motif_barplot(results_subset, group = "class") +
    ggplot2::labs(title = "Retina-only upregulated promoters")
```

# Checking the result against a second gene set

Two very different blocks share the top of that table: the GC-rich zinc fingers,
and the interferon factors the study was about. Nothing in the table tells them
apart. Both are strongly enriched, both are significant by any threshold, and
sorting by z-score puts the zinc fingers first.

The table cannot separate them, because it only ever asks one question. A
z-score says a motif scores higher in this promoter set than across all
promoters of the genome. It does not say whether the motif has anything to do
with the contrast that produced the gene list, and a motif can score high for
reasons that have nothing to do with the tissue or the condition under study.

Scanning a second gene set puts that question directly. Set `f` comes from the
same  same differential expression pipeline; only the tissue differs. 
A motif whose enrichment tracks the contrast should fall away in `f`; 
a motif enriched for any reason the two sets share should not.

```{r second-set-table}
enrichment_f <- precomputed$enrichment_f

data.frame(
    set = c("d", "f"),
    definition = unname(subset_labels[c("d", "f")]),
    promoters = precomputed$subset_qc$scan_eligible_sequences[
        match(c("d", "f"), precomputed$subset_qc$subset)
    ],
    nominal = c(sum(enrichment$P.VALUE < 0.05), sum(enrichment_f$P.VALUE < 0.05)),
    fdr = c(sum(enrichment$FDR < 0.05), sum(enrichment_f$FDR < 0.05))
)
```

Running the same motifs through the striatum set makes the split visible
directly:

```{r second-set-barplot, fig.width=8, fig.height=5, dpi=120}
ps_motif_barplot(precomputed$results_subset_f, group = "class")
```

The GC-rich zinc fingers survive the change of tissue: KLF15 goes from
`r sprintf("%.2f", enrichment["MA1513.1", "ZSCORE"])` to
`r sprintf("%.2f", enrichment_f["MA1513.1", "ZSCORE"])`, still comfortably significant.
The interferon factors do not: IRF3 goes from
`r sprintf("%.2f", enrichment["MA1418.1", "ZSCORE"])` to
`r sprintf("%.2f", enrichment_f["MA1418.1", "ZSCORE"])`, and every IRF and STAT motif
in the panel crosses to the depleted side.

That answers the question the first table could not. The interferon signal
tracks the tissue, and is the result the study reported; the zinc-finger signal
does not, and is shared by upregulated promoters in both tissues. Neither could
have been told from the other by its p-value, however small.

What the comparison does *not* say is why the shared signal is shared. Motifs
common to two gene sets from related tissues can reflect regulators genuinely
active in both, or properties of the promoters that any differential expression
analysis tends to return, and telling those apart takes more than two sets. The
useful conclusion here is the narrower one: only part of this result moves with
the condition, and it is not the part at the top of the table.

## Reproducing the published figure

The published analysis makes the same comparison over all six sets and all
eighteen IRF and STAT matrices, as its Supplementary Figure S2C. 

`ps_results_table()` returns an ordinary `data.frame`, so the six tables stack
into one long frame and anything that draws a heatmap can take it from there.

```{r irf-stat-data}
irf_stat <- precomputed$irf_stat
str(irf_stat[c("subset", "motif", "ZSCORE", "significance")])
```

`motif` is ordered as the published figure orders it, and `significance` marks
the paper's nominal thresholds rather than the FDR, so the two panels can be
laid beside each other.

```{r irf-stat-heatmap, fig.width=10, fig.height=6.7, dpi=120}
ggplot2::ggplot(
    irf_stat,
    ggplot2::aes(x = subset, y = motif, fill = ZSCORE)
) +
    ggplot2::geom_tile(color = "white", linewidth = 0.25) +
    ggplot2::geom_text(ggplot2::aes(label = significance), size = 2.5) +
    ggplot2::scale_fill_gradient2(
        low = "#2166ac", mid = "white", high = "#b2182b",
        midpoint = 0, name = "Z-score"
    ) +
    ggplot2::labs(
        x = NULL, y = NULL,
        title = "IRF and STAT motif enrichment across the six gene sets",
        subtitle = "* nominal P < 0.05; ** nominal P < 0.005"
    ) +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(
        panel.grid = ggplot2::element_blank(),
        axis.text.x = ggplot2::element_text(angle = 35, hjust = 1),
        plot.title.position = "plot"
    )
```

In the retina-only upregulated set 17 of the 18 matrices have a positive z-score 
and 12 pass FDR; in the three downregulated sets the family is *depleted* 
on average, most strongly in the striatum-only one. 
That is a stronger statement than the `d`-against-`f`
comparison above, because it shows the signal is absent not just from the other
tissue but from the other direction of change as well.

```{r irf-stat-counts, echo=FALSE}
counts <- do.call(rbind, lapply(split(irf_stat, irf_stat$subset), function(x) {
    data.frame(
        set = as.character(x$subset[[1]]),
        `nominal P < 0.05` = sum(x$P.VALUE < 0.05),
        `FDR < 0.05` = sum(x$FDR < 0.05),
        `mean Z` = round(mean(x$ZSCORE), 2),
        check.names = FALSE
    )
}))
knitr::kable(
    counts, row.names = FALSE,
    caption = paste(
        "The eighteen IRF and STAT matrices, counted in each set."
    )
)
```

Set `e` is the exception in both columns, but it rests on
`r precomputed$subset_qc$scan_eligible_sequences[precomputed$subset_qc$subset == "e"]`
promoters, too few to read in either direction, which is why the paper's own
contrast is `d` against `f`.

# What a result object contains

`ps_results_table()` summarises across promoters, but the `PSMatrixList` keeps
the per-promoter detail and can be explored with the accessors.

A `PSMatrixList` is indexed by JASPAR matrix identifier, so `MA1418.1` pulls out
the entry for IRF3:

```{r accessors}
irf3 <- results_subset[["MA1418.1"]]
TFBSTools::name(irf3)

ps_zscore(irf3)
ps_pvalue(irf3)
c(background = ps_bg_avg(irf3), foreground = ps_fg_avg(irf3))
ps_hits_size(irf3)
```

Note how small the difference in mean score is that produces a z-score of
`r sprintf("%.2f", ps_zscore(irf3))`. The z-score is a statement about a *mean* over
`r ps_hits_size(irf3)` promoters, not about the size of the effect in any one of
them.

`ps_hits_table()` returns the best hit in every promoter: its score, position,
strand, and the matched oligonucleotide.

```{r hits-table}
head(ps_hits_table(irf3, pos_shift = -950L), 5)
```

The oligo is always the **forward** strand of the promoter, 
whatever strand the motif matched on. The IRF3 matrix is a tandem `GAAAC`
repeat, an interferon-stimulated response element:

```{r consensus}
pfm <- TFBSTools::Matrix(irf3)
paste(rownames(pfm)[apply(pfm, 2, which.max)], collapse = "")
```


# Structure across promoters

`ps_z_table()` gives the per-promoter z-scores as a matrix.

```{r z-table}
z <- ps_z_table(results_subset)
dim(z)
round(z[1:4, 1:5], 2)
```

`ps_zscore_heatmap()` clusters that matrix, which exposes how motifs group
across the promoter set.

```{r zscore-heatmap, fig.width=9, fig.height=7, dpi=110}
ps_zscore_heatmap(
    results_subset, FDR = 1,
    show_rownames = FALSE, treeheight_row = 0
)
```

`FDR = 1` displays every motif supplied. The motifs here were already chosen
from the full 746-motif scan, so there is nothing left to filter.

The column dendrogram recovers the same split, this time from the per-promoter
scores alone and without reference to the second gene set:
the three C2H2 zinc fingers cluster on one branch, everything else on the other.
Promoters scoring high across the zinc-finger columns are largely not the ones
scoring high across the interferon columns. The enrichment table reports both
blocks as significant; only the per-promoter view shows that they are carried by
different promoters.

# Where the sites are

`ps_hitpos_map()` plots the position of the best hit in each promoter.

```{r hitpos-map, fig.width=9, fig.height=7, dpi=110}
ps_hitpos_map(
    results_subset, FDR = 1, shift = -950,
    show_rownames = FALSE, treeheight_row = 0
)
```

`shift = -950` relabels the axis so that coordinates are relative to the
transcription start site, matching how the promoters were extracted. The
zinc fingers separate from the rest again, and here the difference between the
branches is plainly positional: their columns are dominated by hits within a
couple of hundred bases of the start site, where the interferon columns run
through the whole window.

For a single motif, `ps_density_plot()` shows the distribution of hit positions
directly.

```{r density-plot, fig.width=7.5, fig.height=4.6, dpi=120}
ps_density_plot(
    irf3,
    shift = -950, st = "strict",
    window = c(-950, 50), bins = 20
)
```

Strong IRF3 sites rise towards the transcription start site rather than being
spread evenly over the kilobase scanned, though the rise is gradual and the
distribution keeps substantial weight far upstream.

`window` defines the promoter window the positions came from. Given the window,
the kernel density estimate is corrected by reflection. The reflecting bounds
are the positions a hit can be *reported* at: a hit is recorded at the start of
the motif, so it can never fall in the last few bases of the window. IRF3 is
`r ncol(irf3)` bp wide, so its curve stops `r ncol(irf3) - 1` bp short of the
downstream end.

`bins` adds the same positions as a binned profile. Bars are on the density
scale, so they share the curve's axis directly, and they carry no bandwidth and
no boundary assumption, which is what makes them worth drawing next to the
curve. Here the last bin reaches about 1.6 times the height of the smoothed
curve beneath it: the default bandwidth is wide enough to flatten a peak that
occupies only the final fiftieth of the window.

The `st` argument sets which hits are used: `"all"` uses every promoter,
`"loose"` those scoring above the motif's background mean, and `"strict"` those
a standard deviation above it. Restricting to strong sites is usually what makes
positional structure visible.

```{r density-counts, include=FALSE}
irf3_scores <- ps_hits_score(irf3)
loose_n <- sum(irf3_scores >= ps_bg_avg(irf3))
strict_n <- sum(irf3_scores >= ps_bg_avg(irf3) + ps_bg_std_dev(irf3))
```

For IRF3 the loose threshold keeps `r loose_n` of the `r ps_hits_size(irf3)`
promoters and the strict one keeps `r strict_n`. That choice is made *after* the
scan and does not touch the enrichment statistic, which is computed over all of
them.

`ps_hit_score_plot()` puts one point per promoter at its hit position and its
hit score, with the two thresholds `st` names drawn as references.

```{r hit-score-plot, fig.width=9.5, fig.height=4.6, dpi=115}
ps_hit_score_plot(
    results_subset[c("MA1418.1", "MA1513.1")],
    shift = -950
)
```

This separates the two things a density plot confounds. Both motifs lean towards
the transcription start site, but KLF15's retained hits also pile up against the
ceiling of the score range in the last two hundred bases, many of them at or
near a perfect match. IRF3's retained hits lean the same way far more weakly and
stay close to the threshold throughout, which is what a modest signal spread
over many promoters looks like. A density plot would have made the two look
alike; position and score together do not.

# Narrowing the question with a filter

The scan so far asked what is enriched across the whole set of retina-only
upregulated promoters. A narrower question is what is enriched *in the subset of
those promoters that carry a strong match for a particular factor*.

`pscan_filtered()` scans the promoters with one chosen matrix, keeps those
scoring at least `n` standard deviations above that motif's background mean, and
rescans the survivors against the whole collection.

The following chunk is not evaluated, to limit the vignette's build time, but it
can be run as it stands.

```{r pscan-filtered, eval=FALSE}
irf3_matrix <- background[[which(TFBSTools::ID(background) == "MA1418.1")]]

filtered <- pscan_filtered(
    retina_up_promoters, irf3_matrix, n = 1,
    background = background,
    BPPARAM = BiocParallel::SerialParam()
)
ps_results_table(filtered)
```

Anchoring on IRF3 keeps `r precomputed$filtered_irf3$n_retained` of the
`r precomputed$filtered_irf3$n_input` retina-only upregulated promoters.

```{r filtered-table, echo=FALSE}
ft <- precomputed$filtered_irf3$table
ft <- head(ft[order(-ft$ZSCORE), ], 12)
knitr::kable(
    data.frame(
        motif_id = ft$motif_id, NAME = ft$NAME,
        ZSCORE = round(ft$ZSCORE, 2),
        P.VALUE = format(ft$P.VALUE, scientific = TRUE, digits = 3)
    ),
    row.names = FALSE, align = c("l", "l", "r", "r"),
    caption = paste(
        "Enrichment within the", precomputed$filtered_irf3$n_retained,
        "promoters carrying a strong IRF3 site."
    )
)
```

The whole interferon family now sits far above everything else, which is the
expected consequence of selecting on one of its members. 
What is informative is the rest of the list: the GC-rich
zinc fingers that dominated the unfiltered scan have dropped out of the top
entirely.

```{r filtered-contrast, echo=FALSE}
zinc <- c("MA1513.1", "MA1650.1", "MA0162.4")
full <- precomputed$filtered_irf3$table
knitr::kable(
    data.frame(
        NAME = enrichment[zinc, "NAME"],
        `Z all promoters` = round(enrichment[zinc, "ZSCORE"], 2),
        `Z IRF3-positive` = round(
            full$ZSCORE[match(zinc, full$motif_id)], 2
        ),
        check.names = FALSE
    ),
    row.names = FALSE,
    caption = "The GC-rich block, before and after anchoring on IRF3."
)
```

Apparently the promoters carrying a strong IRF3 site are a different population
from the ones driving the GC-rich signal. This is the same conclusion the
score heatmap and the position map pointed at, arrived at independently.

That is the point of the filter: it separates a mixed foreground into
subpopulations that a single scan reports only in aggregate.

# Summary

A complete PscanR analysis is four steps: choose a background, resolve genes to
promoters with a matching window, scan, and read the table.

```{r workflow, eval=FALSE}
background <- generate_psmatrixlist_from_background(
    "Jaspar2020", "mm", c(-950, 50), "mm10", version = 2
)
retina_up_promoters <- ps_select_promoters(
    genes = retina_up_genes, promoter_sequences = promoters_mm10,
    annotation = annotation, gene_col = "ENSEMBL", transcript_col = "REFSEQ",
    mode = "representative", return = "sequences"
)
results <- pscan(retina_up_promoters, background)
ps_results_table(results)
```

Everything after that is interpretation: per-promoter scores (`ps_z_table()`),
hit positions and sequences (`ps_hits_table()`, `ps_hitpos_map()`,
`ps_density_plot()`, `ps_hit_score_plot()`), and the ability to restrict the
foreground to promoters carrying a particular site (`pscan_filtered()`).

# Appendix: how this compares with the published analysis

The study used here ran the original Pscan implementation itself, 
so the two analyses can be compared. The motif-level agreement is shown above, 
in the reproduction of Supplementary Figure S2C: all nine IRF matrices in 
JASPAR 2020 pass FDR 0.05 in the retina-only upregulated set, together
with `STAT1::STAT2`, `Stat2` and `Stat4`, and none passes FDR anywhere else. 

Two differences in setup are worth naming.

```{r selection-counts, include=FALSE}
retina_selection <- ps_selection_summary(retina_mapping)$overall
```

**Transcript selection.** The published analysis does not document which
transcript was used for a gene with several. `ps_select_promoters()` makes that
choice explicit and deterministic, and records it: of the
`r retina_selection$genes_mapped` genes placed here,
`r retina_selection$genes_with_choice` had more than one candidate. Exact motif
ranks may therefore differ from the original web-server run.

**Multiple testing.** PscanR reports the raw *P*-value alongside a
Benjamini–Hochberg FDR; the published figure marks nominal thresholds. Since the
raw *P*-value travels with the result table, any correction can be applied to it
afterwards.

# Session information

```{r session-info}
sessionInfo()
```
